# Multiple Random Variables and Probability Functions

## Multiple random variables

Random variables do not exist in isolation.  We started with a single
random variable $Y$ representing the result of a single, specific coin
flip.  Suppose we fairly flip the coin three times?  Then we can have
random variables $Y_1, Y_2, Y_3$ representing the results of
each of the flips.  We can assume each flip is independent in that it
doesn't depend on the result of other flips.  Each of these variables
$Y_n$ for $n \in 1:3$ has $\mbox{Pr}[Y_n = 1] = 0.5$ and
$\mbox{Pr}[Y_n = 0] = 0.5$.

We can combine multiple random variables using arithmetic operations.
We have already seen comparison operators in writing the event $Y =
1$.  If $Y_1, \ldots, Y_{10}$ are random variables representing ten
coin flips, then we can define their sum as

$$
Z = Y_1 + Y_2 + Y_3
$$

We can simulate values of $Z$ by simulating values of $Y_1, Y_2, Y_3$ and
adding them.

```
y1 = uniform_01_rng()
y2 = uniform_01_rng()
y3 = uniform_01_rng()
z = y1 + y2 + y3
print 'z = ' z
```

It is easier and less error prone to collapse similar values into
arrays and operate on the arrays collectively, or with loops if
necessary.

```
for (n in 1:3)
  y[n] = uniform_01_rng()
z = sum(y)
print 'z = ' z
```

Running this program a few times we get

```{r}
set.seed(1234)
sim_z <- function() {
  y = rbinom(3, 1, 0.5)
  z = sum(y)
  printf('z = %d\n', z)
}
for (k in 1:5) sim_z()
```

We can use simulation to evaluate the probability of an outcome that
combines multiple random variables. For example, to evaluate
$\mbox{Pr}[Z = 2]$, we run the simulation many times and count the
proportion of results that are two.^[The sum is calculated using
notation `sum(y[m, ])`, which is defined to be `sum(y[m, ]) = y[m, 1]
+ ... + y[m, N],` where `N` is the number of entries in row `m` of the
variable `y`.]

```
for (m in 1:M)
  for (n in 1:3)
    y[m, n] = uniform_01_rng()
  z[m] = sum(y[m, ])
Pr_is_two = sum(z == 2) / M
```

As in our other probability estimates, we simulate the variable of
interest $Z$ a total of $M$ times, yielding $z^{(1)}, \ldots,
z^{(m)}$.  Here, that requires simulating $y_1^{(m)}, y_2^{(m)},
y_3^{(m)}$ and adding them for each $z^{(m)}$. We then just count the
number of times `Z` is simulated to be equal to 2 and divide by the
number of simulations.

Letting $M = 100\,000$, and running five times, we get

```{r}
set.seed(1234)
for (k in 1:5) {
  M <- 100000
  is_two <- 0
  for (m in 1:M)
    is_two <- is_two + (rbinom(1, 3, 0.5) == 2)
  printf("Pr[Z == 2] = %4.3f\n", sum(is_two) / M)
}
```

Nailing down that final digit is going to require one hundred times as
many iterations (i.e, $M = 10\,000\,000$ iterations).  Let's see what
that looks like.

```{r}
M <- 10000000
for (k in 1:5)
  printf("Pr[Z == 2] = %4.3f\n", sum(rbinom(M, 3, 0.5) == 2) / M)
```

We can do the same for the other numbers, to get a complete picture of
taking the probability of each number of heads in separate coin flips.

```{r}
M <- 10000000
for (z in 0:3)
  printf("Pr[Z == %d] = %4.3f\n",
         z,
         sum(rbinom(M, 3, 0.5) == z) / M)
```

What if we flip four coins instead of three?

```{r}
M <- 10000000
for (z in 0:4)
  printf("Pr[Z == %d] = %4.3f\n",
         z,
         sum(rbinom(M, 4, 0.5) == z) / M)
```

## Discrete random variables

So far, we have only considered random numbers that take a finite
number of integer values.  A random variable that only takes values in
the integers, i.e., values in

$$
\mathbb{Z} = \ldots -2, -1, 0, 1, 2, \ldots
$$

is said to be a *discrete random variable.*^[In general, any countable
set of numerical values could be used as values of a discrete random
variable.  A set of values is *countable* if each of its members can
be assigned a unique counting number in $\mathbb{N} = 0, 1, 2,
\ldots$.  The integers $\mathbb{Z}$ can be mapped to natural numbers
$\mathbb{N}$ by interleaving, $$\begin{array}{rcl}\mathbb{Z} & &
\mathbb{N} \\ \hline 0 & \mapsto & 0 \\
-1 & \mapsto & 1 \\ 1 & \mapsto & 2 \\ -2 & \mapsto &3 \\ 2 & \mapsto
& 4 \\ & \vdots & \end{array}$$]

## Probability mass functions

It's going to be convenient to have a function that maps each possible
outcome in a variable to its probability.  In general, this will be
possible if and only if the variable is discrete, as defined in the
previous section.

For example, if we reconsider $Z = Y_1 + \cdots Y_4$, the number of
heads in four separate coin flips, we can define a function^[We
implicitly assume that functions return zero for arguments not listed.]

$$
\begin{array}{rclll}
p_Z(0) & = & 1/16 & & \mathrm{TTTT}
\\
p_Z(1) & = & 4/16 & & \mathrm{HTTT, THTT, TTHT, TTTH}
\\
p_Z(2) & = & 6/16 & & \mathrm{HHTT, HTHT, HTTH, THHT, THTH, TTTH}
\\
p_Z(3) & = & 4/16 & & \mathrm{HHHT, HHTH, HTHH, THHH}
\\
p_Z(4) & = & 1/16 & & \mathrm{HHHH}
\end{array}
$$

There are sixteen possible outcomes of flipping four coins.  Because
the flips are separate and fair, each possible outcome is equally
likely.  The sequences corresponding to each count of heads (i.e.,
value of $Z$) are recorded in the rightmost columns.  The
probabilities are derived by dividing the number of ways a value for
$Z$ can arise by the number of possible outcomes.

This function $p_Z$ was constructed to map a value $u$ for $Z$ to the
event probability that $Z = u$,^[Conventionally, this is written as
$$p_Z(z) = \mbox{Pr}[Z = z],$$ but that can be confusing with upper
case $Z$ denoting a random variable and lower case $z$ denoting an
ordinary variable.]

$$
p_Z(u) = \mbox{Pr}[Z = u].
$$

A function defined as above is said to be the *probability mass
function* of the random variable $Z$.  Every discrete random variable has
a unique probability mass function.

Probablity mass functions represent probabilities of a discrete set of
outcomes.  The sum of all such probabilities must be one because at
least one of the outcomes must occur.^[More formally, if $Y$ is a
discrete random variable, then $$\sum_u \, p_Y(u) = 1,$$ where the
summation variable $u$ ranges over all possible values of $Y$.  We are
going to start writing this with the standard overloading of lower and
upper case $Y$ as $$\sum_y \, p_Y(y) = 1.$$]

With large numbers of counts based on simulation, we can more readily
apprehend what is going on with a plot.  Discrete simulations are
typically plotted using bar plots, where the outcomes are arrayed on
the $x$ axis with a vertical bar over each one whose height is
proportional to the frequency of that outcome.

```{r fig.cap="Plot of $M = 100\\,000$ simulations of the probability mass function of a random variable defined as the number of heads in ten specific coin flips."}

set.seed(1234)
M <- 100000
u <- rbinom(M, 10, 0.5)
x <- 0:10
y <- rep(NA, 11)
for (n in 0:10)
  y[n + 1] <- sum(u == n)
bar_plot <-
  ggplot(data.frame(Z = x, count = y), aes(x = Z, y = count)) +
  geom_bar(stat = "identity", colour = "black", fill="#F8F8F0") +
  scale_y_continuous(breaks = c(0, 10000, 20000, 30000), labels =
  c(0, 10000, 20000, 30000)) +
  scale_x_continuous(breaks = c(0, 2, 4, 6, 8, 10),
                     labels = c(0, 2, 4, 6, 8, 10)) +
  ggtheme_tufte()
bar_plot
```

The actual frequencies are not relevant, only the relative sizes.
A simple probability estimate from simulation provides a probability
for each outcome proportional to its height.^[And proportional to its
area because the bars are of equal width.]

This plot can easily be repeated to see what happens as the number of
bins grows.

```{r out.width='80%', fig.cap="Plot of $M = 1\\,000\\,000$ simulations of a variable $Z$ representing the number of heads in $N$ coin flips.  Each plot represents a different $N$.  Because the bars are the same width and the $x$ axes are scaled to the same range in all plots, the total length of all bars laid end to end is the same in each plot;  similarly, the total area of the bars in each plot is the same."}

set.seed(1234)
df <- data.frame()
for (N in c(5, 10, 15, 20, 25, 30)) {
  M <- 1000000
  u <- rbinom(M, N, 0.5)
  x <- 0:N
  y <- rep(NA, N + 1)
  for (n in 0:N)
    y[n + 1] <- sum(u == n)
  df <- rbind(df, list(N = rep(N, N + 1), Z = x, count = y))
}
bar_plot <-
  ggplot(df, aes(x = Z, y = count)) +
  geom_bar(stat = "identity", colour = "black", fill = "#F8F8F0") +
  facet_wrap("N", labeller = labeller(N = label_both)) +
  scale_x_continuous(breaks = c(0, 10, 20, 30), labels = c(0, 10, 20, 30)) +
  scale_y_continuous(breaks = c(), labels = c()) +
  ggtheme_tufte() +
  theme(panel.spacing.x = unit(4, "lines")) +
  theme(panel.spacing.y = unit(2, "lines"))
bar_plot
```

## Dice

Let $Y$ be a random variable representing a fair throw of a six-sided
die. We can describe this variable easily through it's probability
mass function, which is uniform (i.e., assigns each possible outcome
the same probability).

$$
p_Y(y) \ = \
\begin{cases}
\frac{1}{6} & \mbox{if} \ y \in 1:6
\\[4pt]
0 & \mbox{otherwise}
\end{cases}
$$

Games like *Monopoly* use a pair of six-sided dice and consider the sum of the results. That is, $Y_1$ and $Y_2$ are fair six-sided die rolls and $Z = Y_1 + Y_2$ is the result. Games like *Dungeons \& Dragons* use a trio of six-sided dice and consider the sum of the results. In that scenario, $Y_1, Y_2, Y_3$ are the results of fair six-sided die rolls and $Z = Y_1 + Y_2 + Y_3$. Dungeons and Dragons also uses four six-sided die of which the best 3 are summed to produce a result. Let's simulate some of these approaches and see what the results look like based on $M = 100\,000$ simulations.^[The simulations are identical to before, only using `1:6` in the range of uniform variables.]

```{r fig.cap="Estimated $p_Y(y)$ for case of $Y$ being the sum of three six-sided dice (3d6) or the sum of the highest three of four six-sided dice (3 of 4d6)."}

M <- 100000
three_d6 <- sample(6, size = M, replace=TRUE) +
            sample(6, size = M, replace=TRUE) +
            sample(6, size = M, replace=TRUE)
three_of_4d6 <- rep(NA, M)
for (m in 1:M)
  three_of_4d6[m] <- sum(sort(sample(6, size = 4, replace=TRUE))[2:4])

tot_3d6 <- rep(NA, 16)
tot_3_of_4d6 <- rep(NA, 16)
for (n in 3:18) {
  tot_3d6[n-2] = sum(three_d6 == n)
  tot_3_of_4d6[n-2] = sum(three_of_4d6 == n)
}

att_dice_df <- data.frame(probability = c(tot_3d6 / M, tot_3_of_4d6 / M),
                          roll = c(3:18, 3:18),
                          dice = c(rep("3d6", 16), rep("3 of 4d6", 16)))
att_dice_plot <-
  ggplot(att_dice_df, aes(x = roll, y = probability)) +
  geom_bar(stat = "identity", colour = "black", fill = "#F8F8F0") +
  facet_wrap("dice", labeller = labeller(dice = label_both)) +
  scale_x_continuous(name = "y",
                     breaks = c(3, 6, 9, 12, 15, 18),
                     labels = c(3, 6, 9, 12, 15, 18)) +
  scale_y_continuous(name = expression(paste("estimated ", p[Y](y)))) +
  ggtheme_tufte() +
  theme(panel.spacing = unit(2, "lines"))
att_dice_plot
```

Dungeons and Dragons also uses 20-sided dice.^[In physical games, an
icosahedral die is used. The icosahedron is a regular polyhedron with
20 equilateral triangular faces, giving it the largest number of faces
among the five Platonic solids.] The fifth edition of the game
introduced the notion of *advantage*, where two 20-sided dice are
rolled and the higher result retained, as well as *disadvantage*,
which retains the lower result of the two dice. Here's a simulation
using $M = 100\,000$. The counts are converted to estimated
probabilities on the vertical axis in the usual way by dividing by
$M$.

```{r fig.cap="Estimated $p_Y(y)$ for case of $Y$ being a single twenty-sided die (d20), the higher two twenty-sided die rolls (max 2d20), and the lower of two 20-sided die rolls (min 2d20)."}

M <- 100000
d20 <- sample(20, size = M, replace=TRUE)
max_2d20 <- rep(NA, M)
min_2d20 <- rep(NA, M)
for (m in 1:M) {
  max_2d20[m] <- max(sample(20, size = 2, replace=TRUE))
  min_2d20[m] <- min(sample(20, size = 2, replace=TRUE))
}

tot_d20 <- rep(NA, 20)
tot_max_2d20 <- rep(NA, 20)
tot_min_2d20 <- rep(NA, 20)

for (n in 1:20) {
  tot_d20[n] = sum(d20 == n)
  tot_max_2d20[n] = sum(max_2d20 == n)
  tot_min_2d20[n] = sum(min_2d20 == n)
}

att_d20_df <- data.frame(probability = c(tot_d20 / M,
                                         tot_max_2d20 / M,
                                         tot_min_2d20 / M),
                          roll = c(1:20, 1:20, 1:20),
                          dice = c(rep("1d20", 20),
                                   rep("max 2d20", 20),
                                   rep("min 2d20", 20)))

att_d20_plot <-
  ggplot(att_d20_df, aes(x = roll, y = probability)) +
  geom_bar(stat = "identity", colour = "black", fill = "#F8F8F0") +
  facet_wrap("dice", labeller = labeller(dice = label_both)) +
  scale_x_continuous(name = "y",
                     breaks = c(1, 5, 10, 15, 20),
                     labels = c(1, 5, 10, 15, 20)) +
  scale_y_continuous(name = expression(paste("estimated ", p[Y](y)))) +
  ggtheme_tufte() +
  theme(panel.spacing = unit(2, "lines"))

att_d20_plot
```



The most likely roll is a 20 when taking the best of two rolls and the
most likely roll is 1 when taking the worst of two rolls.^[The chance
for a 20 when taking the best of two 20-sided die rolls is $1 -
\left(\frac{19}{20}\right)^2 \approx 0.098$; the chance of rolling 1
is $\left(\frac{1}{20}\right)^2 = 0.0025$. The probabilities are
reversed when taking the worst of two 20-sided die rolls.] The min and
max plots are mirror images of each other as is to be expected by the
consecutive nature of the numbers and the min/max operations.

## Spinners and the Bernoulli distribution

Some games, such as *All Star Baseball*, come with spinners rather
than dice.  The beauty of spinners is that they can be divided into
two areas, one with a 27% chance of occurring in a fair spin and one
with a 73% chance of occurring.^[27% is roughly the chance of a hit in
an at bat.]  Now suppose we have a random variable $Y$ representing
the result of a fair spin.  Its probability mass function is

$$
p_Y(y)
\ = \
\begin{cases}
0.27 & \mbox{if} \ \ y = 1
\\
0.73 & \mbox{if} \ \ y = 0
\end{cases}
$$

To simplify our notation, we are going to start defining useful
functions that can be used as probability mass functions.  Our first
example is the so-called Bernoulli^[Named after Jacob Bernoulli
(1654--1705), one of several prominent mathematicians in the family.]
distribution.  We define it as a function with somewhat peculiar
notation,

$$
\mathrm{Bernoulli}(y \mid \theta)
\ = \
\begin{cases}
\theta & \mbox{if} \ \ y = 1
\\
1 - \theta & \mbox{if} \ \ y = 0
\end{cases}
$$

The vertical bar ($\mid$) separates the *variate* argument $y$, which
we think of as an outcome, from the *parameter* argument $\theta \in
[0, 1]$, which determines the probability of the outcome.  In this
case, the variate $y$ is discrete, and can take on only the values
zero and one, so we write $y \in 0:1$.  The parameter $\theta$, on the
other hand, is continuous and can take on any value between zero and
one (inclusive of endpoints), so we write $y \in [0, 1]$.^[Interval
notation $[0, 1]$ is used for the set of values $x$ such that $0 \leq
x \leq 1$.  Parentheses are used for exclusive endpoints, so that $(0,
1)$ is taken to be the set of $x$ such that $0 < x < 1$.]

This notation allows us to simplify our baseball example.  Going back
to our example random variable $Y$ which had a 27% chance of being 1
and a 73% chance of being 0, we can write

$$
p_Y(y) = \mathrm{Bernoulli}(y \mid 0.27).
$$

To simplify notation even further, we will say that a random variable
$U$ has a Bernoulli distribution and write

$$
U \sim \mathrm{Bernoulli}(\theta)
$$

to indicate that the probability mass function of $U$ is

$$
p_U(u) = \mathrm{Bernoulli}(u \mid \theta).
$$


## Cumulative distribution functions

In Dungeons \& Dragons, the players are often concerned with
probabilities of rolling higher than a given number (or equivalently,
rolling lower than a given number. For example, they may need to roll
a 15 to sneak by an orc. Such probabilities are conventionally given
in the form of cumulative distribution functions. If $Y$ is a random
variable, its *cumulative distribution function* $F_Y$ is defined by

$$
F_Y(y) = \mbox{Pr}[Y \leq y].
$$

The event probability on the right is calculated the same way as
always in a simulation, by counting the number of simulated values in
which the condition holds and dividing by the number of simulations.

We can plot the cumulative distribution function for the straight
twenty-sided die roll and the rolls with advantage (best of two rolls)
or disadvantage (worst of two rolls).  Here's the result using the
same simulations as in the last plot, with $M = 100\,000$.

```{r fig.cap="Cumulative distribution function for three variables corresponding to rolling a single 20-sided die, or rolling two 20-sided dice and taking the best or worst result."}
cum_d20 <- cumsum(tot_d20)
cum_max_2d20 <- cumsum(tot_max_2d20)
cum_min_2d20 <- cumsum(tot_min_2d20)

cum_d20_df <- data.frame(probability = c(cum_d20 / M,
                                         cum_max_2d20 / M,
                                         cum_min_2d20 / M),
                          roll = c(1:20, 1:20, 1:20),
                          dice = c(rep("1d20", 20),
                                   rep("max 2d20", 20),
                                   rep("min 2d20", 20)))

cum_d20_plot <-
  ggplot(cum_d20_df, aes(x = roll, y = probability)) +
  geom_point() +
  facet_wrap("dice", labeller = labeller(dice = label_both)) +
  scale_x_continuous(name = "y",
                     breaks = c(1, 5, 10, 15, 20),
                     labels = c(1, 5, 10, 15, 20)) +
  scale_y_continuous(name = expression(paste("estimated ", F[Y](y)))) +
  ggtheme_tufte() +
  theme(panel.spacing = unit(2, "lines"))
cum_d20_plot
```

** TODO(carpenter): get the right CDF plot and fix aspect ratio. **

The plot is rendered as a point plot, though this isn't quite sensible
for discrete distributions---the intermediate values are not real.
It's easy to see that there's a 50% chance of rolling 5 or lower in a
single die throw; with the best of 2 it's more like a 20% chance and
with the worst of 2, it's more like a 75% chance.

Usually in Dungeons \& Dragons, players care about rolling more than a
given number, not less, or they'd have to be subtracting all the time.
This is where the complementary cumulative distribution function comes
in.  For a random variable $Y$, the *complementary cumulative
distribution function* is

$$
F^{\complement}_Y(y) \ = \ 1 - F_Y(y) \ = \ \mbox{Pr}[Y > y].
$$

It's easier to see with a plot how it relates to the usual cumulative
distribution function.

```{r, fig.cap = "Complementary cumulative distributions for a single 20-sided die, the best of two dice, and the worst of two dice."}
ccum_d20_df <- data.frame(probability = c(1 - cum_d20 / M,
                                          1 - cum_max_2d20 / M,
                                          1 - cum_min_2d20 / M),
                          roll = c(1:20, 1:20, 1:20),
                          dice = c(rep("1d20", 20),
                                   rep("max 2d20", 20),
                                   rep("min 2d20", 20)))

ccum_d20_plot <-
  ggplot(ccum_d20_df, aes(x = roll, y = probability)) +
  geom_point() +
  facet_wrap("dice", labeller = labeller(dice = label_both)) +
  scale_x_continuous(name = "y",
                     breaks = c(1, 5, 10, 15, 20),
                     labels = c(1, 5, 10, 15, 20)) +
  scale_y_continuous(name = expression(paste("estimated ",
                                             F[Y]^"C", "(y)"))) +
  ggtheme_tufte() +
  theme(panel.spacing = unit(2, "lines"))
ccum_d20_plot
```



## Infinite discrete random variables

Consider an experiment in which a coin is tossed until a heads appears
Let the random variable $U$ be the number of tosses that came up tails
before the first head comes up.  The legal sequences are H (0 tails),
TH (1 tails), TTH (2 tails), and so on.  There is no upper limit to
how many tails may appear before the first heads.

Here's some code to create $M$ simulations of the variable $U$.^[This
code uses a while loop, which repeats as long as its condition
evaluates to true (i.e., 1).  Here, the condition compares the output
of the random number generator directly rather than assigning to an
intermediate value.  We have also introduced the increment operator
`+=`, which adds the value of the right hand side to the variable on
the left hand side.]

```
for (m in 1:M)
  u[m] = 0
  while (uniform_01_rng() == 0)
    u[m] += 1
```

This looks dangerous!  The body of a while-loop (here `u[m] + 1`) is
executed iteratively as long as the condition is true.^[If we write
`while (1 + 1 == 2)` we produce what is known as an *infinite loop*,
i.e., one that never terminates.]  Shouldn't we be worried that the
random number generator will just continue to throw tails (i.e., 0)
so that the program never terminates?^[The answer is "yes," in general,
because programmers are error prone.]  In this case, no, because the
odds are vanishingly small that $U$ gets large.  For example,

$$
\begin{array}{rcl}
\mbox{Pr}[U < 10]
& = &
p_U(0) + p_U(1) + \cdots p_U(9)
\\[6pt]
& = &
\frac{1}{2} + \cdots + \frac{1}{1024}
\\[6pt]
& \approx &
0.999.
\end{array}
$$

Going further, $\mbox{Pr}[U < 20] \approx 0.999\,999$, and so on.  So
there's not much chance of running very long at all, much less forever.

With the concern of non-termination out of the way, let's see what we
get with $M = 50$ simulations of $U$.

```{r}
sim_u <- function() {
  u <- 0
  while (rbinom(1, 1, 0.5) == 0)
    u <- u + 1
  u
}
set.seed(1234)
for (m1 in 1:5) {
  for (m2 in 1:10)
    printf("%4d", sim_u())
  printf("\n")
}
```

It's very hard to discern a pattern here.  There are a lot of zero
values, but also some large values.  For cases like these, we can use
a bar plot to plot the values.  This time, we're going to use $M =
10\,000$ to get a better picture of the pattern.

```{r fig.cap="Frequency of outcomes in $10\\,000$ simulation draws of $U$, the number of tails seen before a head in a coin-tossing experiment."}
set.seed(1234)
M <- 10000
u <- rep(NA, M)
for (m in 1:M)
  u[m] <- sim_u()
x <- 0:max(u)
y <- rep(NA, max(u) + 1)
for (n in 0:max(u))
  y[n + 1] <- sum(u == n)
bar_plot <-
  ggplot(data.frame(U = x, count = y), aes(x = U, y = count)) +
  geom_bar(stat = "identity", colour = "black", fill = "#F8F8F0") +
  scale_x_continuous(breaks = c(0:12), labels = c(0:12)) +
  scale_y_continuous(breaks = c(1000, 2000, 3000, 4000, 5000),
                     labels = c(1000, 2000, 3000, 4000, 5000)) +
  ggtheme_tufte()
bar_plot
```

The $x$-axis represents the value of $U$ and the $y$-axis the number
of times that value arose in the simulation.^[Despite $U$ having
infinitely many possible values, it will only take on finitely many of
them in a finite sample.]  Each additional throw of tails appears to
cut the probability of occurrence in half result seems to have about
half the probability of the previous one.  This is what we should
expect because each coin toss brings a 50% probability of a tails
result.  This exponential decay^[Exponential decay means each
additional outcome is only a fraction as likely as the previous one.]
in the counts with the number of tails thrown is more obvious when
plotted on the log scale.

```{r fig.cap="Frequency of outcomes in $10\\,000$ simulation draws of $U$, the number of tails seen before a head in a coin-tossing experiment, this time with the outcome count on the log scale to illustrate the exponentially decreasing probabilities of each successive number of tails."}

set.seed(1234)
M <- 10000
u <- rep(NA, M)
for (m in 1:M)
  u[m] <- sim_u()
x <- 0:max(u)
y <- rep(NA, max(u) + 1)
for (n in 0:max(u))
  y[n + 1] <- sum(u == n)
log_bar_plot <-
  ggplot(data.frame(U = x, count = y), aes(U, count)) +
  geom_bar(stat = "identity", colour = "black", fill = "#F8F8F0") +
  scale_x_continuous(breaks = c(0:12), labels = c(0:12)) +
  scale_y_log10(breaks = c(8, 40, 200, 1000, 5000),
                labels = c(8, 40, 200, 1000, 5000)) +
  ggtheme_tufte()
log_bar_plot
```

There is a 50% probability that the first toss is heads, yielding a
sequence of zero tails, and $U = 0$.  Each successive number of tails
is half as likely as the previous, because another tail will have to
be thrown, which has a 50% probability.^[In symbols,
$$
\mbox{Pr}[U = n + 1]
\ = \
\frac{1}{2} \mbox{Pr}[U = n].
$$
]

Thus the overall probability mass function for $U$ is^[An elementary
result of calculus is that $$\sum_{n = 0}^{\infty}
\frac{1}{2^{n + 1}} \ = \ \frac{1}{2} + \frac{1}{4} + \frac{1}{8} +
\frac{1}{16} + \cdots \ = \ 1.$$]

$$
\begin{array}{rclcl}
p_U(0) & = & \frac{1}{2} & = & \frac{1}{2}
\\
p_U(1) & = &  \frac{1}{2} \times \frac{1}{2} & = & \frac{1}{4}
\\
p_U(2) & = & \frac{1}{2} \times \frac{1}{2} \times \frac{1}{2} & = & \frac{1}{8}
\\
& \vdots &
\\
p_U(u) & = & \underbrace{\frac{1}{2} \times
                         \cdots \times \frac{1}{2}}_{u + 1 \ \mathrm{times}}
& = &
\left( \frac{1}{2} \right)^{u + 1}
\\
& \vdots &
\end{array}
$$

Even though there are infinitely many possible realizations of the
random variable $U$, simulation may still be used to compute event
probabilities, such as $\mbox{Pr}[U \leq 3]$, by

```
for (m in 1:M)
  u[m] = sim_u()
  leq3[m] = (u[m] <= 3)

print 'est Pr[U <= 3] = ', sum(leq3) / M
```

Let's see what we get with $M = $100\,000$,

```{r}
set.seed(1234)
M <- 100000
sum_leq3 <- 0
for (m in 1:M) {
  u_m <- sim_u()
  if (u_m <= 3) sum_leq3 <- sum_leq3 + 1
}
printf('est Pr[U <= 3] = %4.3f\n', sum_leq3 / M)
```

Writing out the analytic answer involves an infinte sum,

$$
\mbox{Pr}[U \leq 3]
\ = \
\sum_{u = 0}^{\infty} p_U(u) \mathrm{I}[u \leq 3].
$$

We can recognize that all of the terms where $u > 3$ are zero, so that
this reduces to

$$
\begin{array}{rcl}
\mbox{Pr}[U \leq 3]
& = &
p_U(0) + p_U(1) + p_U(2) + p_U(3)
\\[8pt]
& = &
\frac{1}{2}
+ \frac{1}{4}
+ \frac{1}{8}
+ \frac{1}{16}
\\[8pt]
& \approx & 0.938.
\end{array}
$$

Simulation is not that clever.  It just blindly simulates values of
$u$, many of which turn out to be larger than three.  In the sequence
of simulated values, many were larger than three---the histogram
summarized a much larger simulation, none of the values of which were
that large.

Given that we only ever run a finite number of iterations and thus
only ever see a finite number of values, how does simulation get the
right answer?^[Spoiler alert!  The technical answer, like so much in
statistics, is that the central limit theorem kicks in.]  Isn't there
some kind of bias from only visiting smallish values?  The answer is
"no" precisely because the values that are not simulated are so rare.
Their total probability mass, when added together is small, so they
cannot have much influence on the simulated answer.  For simulations
to get the right answer,^[Given some qualifications!] they need only
visit the typical values seen in simulation, not the extreme
values.^[The sparsity problem grows exponentially worse in higher
dimensions and uncountably worse with continuous random variables.]


## The Chevalier de Méré's challenge

Antoine Gombaud, the Chevalier de Méré,^[Self appointed!] challenged
Blaise Pascal to explain how it was possible that thre probability of
throwing at least one six in four throws of a single six-sided die is
slightly greater than $\frac{1}{2}$, whereas the probability of
throwing two sixes in 24 throws of a pair of six-sided die was
slightly less than $\frac{1}{2}$.^[Smith, D. A., 1929.  *A Source Book
on Mathematics*. McGraw-Hill, New York; cited in Bulmer, 1967, p. 26.]
We can evaluate these claims by simulation directly.^[Working out the
example analytically, there is a $\frac{35}{36}$ chance of *not*
throwing double six with two six-sided dice, and so $\left(
\frac{35}{36} \right)^{24}$ is the probability of *not* throwing at
least one double six in 24 throws, and so $$1 - \left( \frac{35}{36}
\right)^{24} \approx 0.491$$ is the probability of throwing at least
one double-six in 24 fair throws of a pair of six-sided dice.]

To represent the problem in probabilistic notation, we introduce a
random variable $Y_{1, k} \in 1:6$ and $Y_{2, k} \in 1:6$ for each of
the two dice in each of the $k \in 1:24$ throws. Define the outcome of
the game as the random variable

$$
Z \ = \
\begin{cases}
1 \ \ \mbox{if there is a} \ k \ \mbox{such that} \ Y_{1,k} = Y_{2,k} = 6,
\ \mbox{and}
\\[4pt]
0 \ \ \mbox{otherwise}
\end{cases}
$$

That is, $Z = 1$ if there is at least one double-six in the 24 throws.
The Chevalier de Méré was inquiring about the the value of the event
probability $\mbox{Pr}[Z = 1]$, i.e., the chance of winning by
throwing at least one double six in 24 fair throws of a pair of dice.

We will introduce variables for the full range of simulated random
variable values and simulation indexes in the following program.^[The
simulation indexes use parentheses rather than the traditional
brackets and come first so that, e.g., the simulated value `y(m)` will
consist of a $2 \times 24$ collection of values, matching the size of
$Y$.]

```
for (m in 1:M)
  for (k in 1:24)
    y(m)[1, k] = uniform_rng(1:6)
    y(m)[2, k] = uniform_rng(1:6)
    z(m)[k] = y(m)[1, k] + y(m)[2, k]
  success(m) = (sum(z(m) == 12) > 0)
print 'Pr[double-six in 24 throws] = ' sum(success) / M
```

Let's run that for $M = 100\,000$ simulations a few times and see
what the estimated event probabilities look like.

```{r}
set.seed(1234)
M <- 1e5
for (k in 1:5) {
  successes = 0
  for (m in 1:M) {
    z <- sample(6, size = 24, replace = TRUE) +
         sample(6, size = 24, replace = TRUE)
    if (sum(z == 12) > 0) successes <- successes + 1
  }
  printf('Pr[double-six in 24 throws] = %4.3f\n', successes / M)
}
```

This shows the result to be around 0.49. The Chevalier de Méré
should not bet that he'll roll at least one pair of sixes in 24
throws! To nail down the last digit, we could use $10\,000\,000$
simulations rather than $100\,000$. As shown in the previous note,
calculating the result analytically yields 0.491 to three decimal
places, which is in agreement with the simulation-based estimates.

The Chevalier de Méré was reputedly perplexed by the difference
between the chance of at least one double-six in 24 throws of two dice
versus the chance of at least one six in 4 throws of a single
die.^[The probability of at least one six in four die rolls works out
to $$1 - \left( \frac{5}{6} \right)^4 \approx 0.518.$$ As noted above,
the probability of at least one double six in 24 die rolls is $\approx
0.491.$]


## Sampling without replacement

Drawing from a deck of cards is typically done without replacement.
Once a card is drawn, it may not be drawn again.  A traditional deck
of playing cards consists of 52 cards, each marked with a value

$$
2, 3, \ldots, 10, \mathrm{J}, \mathrm{Q}, \mathrm{K}, \mathrm{A}
$$

and a suit from

$$
\clubsuit, \, \diamondsuit, \ \heartsuit, \ \spadesuit
$$

The lettered values are called jack, queen, king, and ace, and the
suits are called clubs, diamonds, hearts, and spades.  Traditionally
the diamonds and hearts are colored red and the clubs and spades
colored black (here they are indicated by unfilled vs.\ filled
shading).

A hand of cards consists of some number of cards drawn from a deck.
When cards are drawn from a deck, they are not replaced.  This is
called sampling without replacement.  Thus a hand of cards can contain
at most 52 cards, because after 52 cards are drawn, there are none
left.^[In some games, multiple decks are often used.]  Drawing without
replacement also affects probabilities of particular hands.

For example, drawing two cards from a fresh deck, the chance of
getting two aces is not $\left(\frac{4}{52}\right)^2$, but rather
$\left(\frac{4}{52} \times \frac{3}{51}\right) \approx 0.0045.$ The
chance of drawing an ace on the first draw is $\frac{4}{52}$ because
there are 4 aces among the 52 cards and each card is assumed to be
equally likely to be drawn from any deck.  But after the first ace is
drawn, there are only 51 cards remaining, and among those, only 3
aces.  So the chance of the second card being an ace is only
$\frac{3}{51}$.

We can verify that with a quick simulation.

```
total = 0
for (m in 1:M)
  y <- draw_cards(2)
  if (is_ace(y[1]) && is_ace(y[2]))
    total += 1
print 'Pr[draw 2 aces] = ' total / M
```

Let's run that with $M = 10\,000$, a few

```{r}
set.seed(1234)
M <- 1e4
for (k in 1:8) {
  total <- 0
  for (m in 1:M)
    total <- total + (sum(sample(52, size = 2) >= 49) == 2)
  printf('Pr[draw 2 aces] = %5.4f\n',  total / M)
}
```

Curiously, we are now not getting a single digit of accuracy, even
with $10\,000$ draws.  What happened?

A fundamental problem with accuracy of simulation-based estimates is
that rare events are hard to estimate with random draws.  If the event
of drawing two aces only has a 0.45% chance (roughly 1 in 200) of
occurring, we need a lot of simulation events to see it often enough
to get a good estimate of even that first digit.  With $10\,000$
draws, the number of two-ace draws we expect to see is about 50 if
they occur at roughly a 1 in 200 hands rate.  We know from prior
experience that estimating a number with only 50 draws is not going to
be very accurate.  So what we need to do is increase the number of
draws.  Let's run that again with $M = 1\,000\,000$ draws.

```{r}
set.seed(1234)
M <- 1e6
for (k in 1:4) {
  total <- 0
  for (m in 1:M)
    total <- total + (sum(sample(52, size = 2) >= 49) == 2)
  printf('Pr[draw 2 aces] = %5.4f\n',  total / M)
}
```

Now with an expected $5\,000$ occurrences of a two-ace hand, we have a
much better handle on the relative accuracy, having nailed down at
least the first digit and gotten close with the second digit.


## Error versus relative error

Suppose we have an estimate $\hat{y}$ for a quantity $y$.  One natural
way to measure the accuracy of the estimate is to consider its
*error*,

$$
\mathrm{err} = \hat{y} - y.
$$

If the estimate is too high, the error will be positive, and if the
estimate is too low, the error will be negative.  The problem with
this standard notion of error arises when the estimand $y$ is very
small or very large.

Now consider an estimand of $y = 0.01$ and an estimate of $\hat{y} =
0.015$.  The error is just $y - \hat{y}$ = 0.005$, which looks small.
But compared to the magnitude of $y$, which is only 0.01, the error
is relatively large.

The *relative error* of an estimate $\hat{y}$ of a quantity $y$ can
be defined relative to the scale of the estimand as

$$
\mathrm{rel\_err} = \frac{\hat{y} - y}{\left| \, y \, \right|}.
$$

This delivers results that are scaled in units of of $y$.  The
relative error for our estimate $\hat{y} = 0.015$ for an estimand $y =
0.01$ has relative error of


$$
\frac{0.015 - 0.01}{\left| 0.01 \right|} = 0.5.
$$

That's a 50% relative error, which now looks quite large compared to
the 0.005 error.^[An estimate of $\hat{y} = 0.015$ has an error of
$-0.005$ and a relative error of $-0.5$, or 50% too low.]

If the sign of error doesn't matter, errors are often reported as
absolute values, i.e., as *absolute error* and *absolute relative
error*.

The central limit theorem provides guarantees only about the error;
to calculate its implications for relative error, the true value of
the estimand must be known.

## The Earl of Yarborough's wager

If an event has a probability of $\theta$, the *odds* of it happening are
given by the function

$$
\mathrm{odds}(\theta) = \frac{\theta}{1 - \theta}.
$$

For example, if there is a 25% chance of an event happening, the odds
of it happening are $\frac{0.25}{1 - 0.25} = \frac{1}{3}$.  In other
words, it's three times as probable that the event does not occur than
that it occurs.  Odds are written as $1:3$ and pronounced "one to
three" rather than being written as $\frac{1}{3}$ and pronounced "one
in three".^[When reporting odds, it is common to report the odds as
"three to one *against*" for an event with a 25% probability.]

The Earl of Yarborough^[Many early developments in probability were
bankrolled by gambling aristocrats.] reputedly laid a thousand to one
odds against drawing a 13-card whist hand that contained no card
higher than a 9.^[Bulmer, 1967, p. 26.]  There is a total of 32 cards
that are 9 or lower (and hence 20 cards 10 or higher). We can use the
same argument to calculate the Earl's odds of drawing his eponymous
hand at

$$
\begin{array}{rcl}
\mbox{Pr}\left[\mbox{draw a Yarborough}\right]
& = & \frac{32}{52} \times \frac{31}{51} \times \cdots \times \frac{20}{40}
\\[6pt]
& = & \prod_{n = 0}^{12} \frac{32 - n}{52 - n}
\\[6pt]
& \approx & 0.00055
\end{array}
$$

The $n$ in the second line represents the number of cards drawn
previously. The true odds are roughly one in 2000, or rounded to the
nearest integer

$$
1:1817 \approx \frac{0.00055}{1 - 0.00055}
$$

Rounded to the nearest integer, the odds are 1817:1 against, so the
Earl should expect to profit from his bet.^[Modulo the fact that one
rare event might ruin him---these market-making schemes require a
large bankroll and many repetitions to avoid ruin.]


## Binomial and repeated binary trials

Suppose we have a sequence of random variables, $V_1, \ldots, V_N$,
each with a Bernoulli distribution $V_n \sim
\mathrm{Bernoulli}(\theta)$.  That is, each $V_n$ takes on the value 1
with a probabilty of $\theta$.

We can think of $V_1, \ldots, V_N$ as $N$ repeated binary trials, each
with a $\theta$ chance of success.^[The term "success" is the
conventional name for the result 1 in an abstract binary trial, with
result 0 being "failure".]  That is, each $V_n$ is a completely
independent trial and each trial has a $\theta$ chance of success. By
independent, we mean that the success of $Y_n$ does not depend on
$Y_{n'}$ if $n \neq n'$.

What can we say about the number of successes in $N$ trials?  Let

$$
\begin{array}{rcl}
Y & = & V_1 + \cdots + V_N
\\[3pt]
& = & \sum_{n = 1}^N V_n,
\end{array}
$$

and the question reduces to what we can say about the random variable
$Y$.  Repeated binary trials come up so often that the distribution of
$Y$ has a name, the *binomial distribution*.  Pascal figured out that
for any number of trials $N \geq 0$, chance of success $\theta \in [0,
1]$, the probability of a total number of successes $y \in 0:N$ is

$$
p_Y(y) = \mathrm{Binomial}(y \mid N, \theta),
$$

where^[The value ${N \choose y}$ is called the *binomial coefficient*
due to its use here, and defined by $${N \choose y} = \frac{N!}{(N -
y)! \times y!}.$$ The value of the *factorial* $m!$ for $m > 0$ is
$$m! = m \times (m - 1) \times (m - 2) \times \cdots 1.$$ The recursive
definition has base case $0! = 1$ and inductive case $(n + 1)! = n
\times n!.$ The postfix factorial operator binds more tightly than
multiplication, so this resolves as $n \times (n!)$. ]

$$
\mathrm{Binomial}(y \mid N, \theta)
\ = \
{N \choose y} \times \theta^y \times (1 - \theta)^{N - y}.
$$

## Variance of the Bernoulli and binomial

If $Y \sim \mathrm{Bernoulli}(\theta)$, then

$$
\begin{array}{rcl}
\mathbb{E}[Y]
& = &
\sum_{y \in 0:1} p_Y(y) \times y
\\[2pt]
& = & \mathrm{Bernoulli}(0 \mid \theta) \times 0
      + \mathrm{Bernoulli}(1 \mid \theta) \times 1
\\[2pt]
& = & (1 - \theta) \times 0 + \theta \times 1
\\[2pt]
& = & \theta.
\end{array}
$$

Plugging this into the formula for variance yields

$$
\begin{array}{rcl}
\mathrm{var}[Y]
& = &
\mathbb{E}\left[ (Y - \mathbb{E}[Y])^2 \right]
\\[2pt]
& = &
\sum_{y \in 0:1} p_Y(y) \times (y - \theta)^2
\\[2pt]
& = & (1 - \theta) \times (0 - \theta)^2
      +  \theta \times (1 - \theta)^2
\\[2pt]
& = &
(1 - \theta) \times \theta^2
+ \theta \times (1 - 2 \times \theta + \theta^2)
\\[2pt]
& = &
\theta^2 - \theta^3
+ \theta - 2 \times \theta^2 + \theta^3
\\[2pt]
& = &
\theta - \theta^2
\\[2pt]
& = &
\theta \times (1 - \theta).
\end{array}
$$


## The multinomial distribution

The binomial distribution is for repeated binary trials---the
multinomial distribution extends the same idea to repeated categorical
trials.  Just as a multiple coin tosses can be represented by a
binomial distribution, multiple die rolls can be represented by
multinomial distributions.


## Counts and the Poisson distribution

Consider two binomial random variables,

$$
Y \sim \mathrm{binomial}(N, \theta),
$$

and

$$
Z \sim \mathrm{binomial}(2 \times N, \frac{1}{2} \theta).
$$

The variable $Z$ has a maximum value that is twice as large of that of
$Y$, yet it has the same expectation,

$$
\mathbb{E}[Y] = \mathbb{E}[Z] = N \times \theta.
$$

Poisson is the limit of the binomial in the sense that

$$
\mathrm{Poisson}(y \mid \lambda)
\ = \
\lim_{N \rightarrow \infty}
\mathrm{Binomial}(y \mid N, \frac{1}{N} \lambda)
$$

We can plot this out in a series of binomials:
binomial(N, lambda / N) for N in ceil(lambda), *= sqrt(10) ...