# Random Variables and Event Probabilities

## Random variables

Let $Y$ be the result of a fair coin flip.  Not a general coin flip,
but a specific instance of flipping a specific coin at a specific
time.  Defined this way, $Y$ is what's known as a *random variable*,
meaning a variable that takes on different values with different
probabilities.^[Random variables are conventionally written using
upper-case letters to distinguish them from ordinary mathematical
variables which are bound to single values and conventionally written
using lower-case letters.]

Probabilities are scaled between 0% and 100% as in natural language.
If a coin flip is fair, there is a 50% chance the coin lands face up
("heads") and a 50% chance it lands face down ("tails").  For
concreteness and ease of analysis, random variables will be restricted
to numerical values.  For the specific coin flip in question, the
random variable $Y$ will take on the value 1 if the coin lands heads
and the value 0 if it lands tails.


## Events and probability

An outcome such as the coin landing heads is called an *event* in
probability theory.  For our purposes, events will be defined as
conditions on random variables.  For example, $Y = 1$ denotes the
event in which our coin flip lands heads. The functional $\mbox{Pr}[\,
\cdot \,]$ defines the probability of an event.  For example, for our
fair coin toss, the probability of the event of the coin landing heads
is written as

$$
\mbox{Pr}[Y = 1] = 0.5.
$$

In order for the flip to be fair, we must have $\mbox{Pr}[Y = 0] =
0.5$, too.  The two events $Y = 1$ and $Y = 0$ are mutually exclusive
in the sense that both of them cannot occur at the same time.  In
probabilistic notation,

$$
\mbox{Pr}[Y = 1 \ \mbox{and} \ Y = 0] = 0.
$$

The events $Y = 1$ and $Y = 0$ are also exhaustive, in the sense that
at least one of them must occur.  In probabilistic notation,

$$
\mbox{Pr}[Y = 1 \ \mbox{or} \ Y = 0] = 1.
$$

In these cases, events are conjoined (with "and") and disjoined (with
"or").  These operations apply in general to events, as does negation.
As an example of negation,

$$
\mbox{Pr}[Y \neq 1] = 0.5.
$$


## Sample spaces and possible worlds

Even though the coin flip will have a specific outcome in the real
world, we consider alternative ways the world could have been.  Thus
even if the coin lands heads $(Y = 1)$, we entertain the possibility
that it could've landed tails $(Y = 0)$.  Such counterfactual
reasoning is the key to understanding probability theory and applied
statistical inference.

An alternative way the world could be, that is, a *possible world*,
will determine the value of every random variable. The collection of
all such possible worlds is called the *sample space*.^[The sample
space conventionally written as $\Omega$, the capitalized form of the
last letter in the Greek alphabet.]  The sample space may be
conceptualized as an urn containing a ball for each possible way the
world can be.  On each ball is written the value of every random
variable.^[Formally, a random variable $X$ can be represented as a
function from the sample space to a real value, i.e., $X:\Omega
\rightarrow \mathbb{R}$. For each possible world $\omega \in \Omega$,
the variable $X$ takes on a specific value $X(\omega) \in
\mathbb{R}$.]

Now consider the event $Y = 0$, in which our coin flip lands tails. In
some worlds, the event occurs (i.e., $0$ is the value recorded for
$Y$) and in others it doesn't.  An event picks out the subset of
worlds in which it occurs.^[Formally, an event is defined by a subset
of the sample space, $E \subseteq \Omega$.]




## Simulating random variables

We are now going to turn our attention to computation, and in
particular, simulation, with which we will use to estimate event
probabilities.

The primitive unit of simulation is a function that acts like a random
number generator.  But we only have computers to work with and they
are deterministic.  At best, we can created so-called *pseudorandom
number generators*.  Pseudorandom number generators, if they are well
coded, produce deterministic streams of output that appear to be
random.^[There is a large literature on pseudorandom number generators
and tests for measurable differences from truly random streams.]

For the time being, we will assume we have a primitive pseudorandom
number generator `uniform_01_rng()`, which behaves roughly like it has
a 50% chance of returning 1 and a 50% chance of returning 0.^[The name
arises because random variables in which every possible outcome is
equally likely are said to be *uniform*.]

Suppose we want to simulate our random variable $Y$.  We can do so by
calling `uniform_01_rng` and noting the answer.



A simple program to generate a realization of a random coin flip,
assign it to an integer variable `y`, and print the result could be
coded as follows.^[Computer programs are presented using a consistent
pseudocode, which provides a sketch of a program that should be
precise enough to be coded in a concrete programming language.  R
implementations of the pseudocode generate the results and are
available in the source code repository for this book.]

```
int y = uniform_01_rng()
print 'y = ' y
```

The variable `y` is declared to be an integer and assigned to the
result of calling the `uniform_01_rng()` function.^[The use of a
lower-case $y$ was not accidental.  The variable $y$ represents an
integer, which is the type of a realization of a random $Y$
representing the outcome of a coin flip.  In code, variables are
written in typewriter font (e.g., `y`), whereas in text they are
written in italics like other mathematical variables (e.g., $y$).]
The print statement outputs the quoted string `y = ` &nbsp; followed
by the value of the variable `y`.  Executing the program might produce
the following output.

```{r}
printf("y = %d", rbinom(1, 1, 0.5))
```

If we run it a nine more times, it might print

```{r}
printf("y = %d", rbinom(1, 1, 0.5))
printf("y = %d", rbinom(1, 1, 0.5))
printf("y = %d", rbinom(1, 1, 0.5))
printf("y = %d", rbinom(1, 1, 0.5))
printf("y = %d", rbinom(1, 1, 0.5))
printf("y = %d", rbinom(1, 1, 0.5))
printf("y = %d", rbinom(1, 1, 0.5))
printf("y = %d", rbinom(1, 1, 0.5))
printf("y = %d", rbinom(1, 1, 0.5))
```
When we say it might print these things, we mean the results will
depend on the state of the pseudorandom number generator.

## Seeding a simulation

Simulations can be made exactly reproducible by setting what is known
as the *seed* of a pseudorandom number generator.  This seed
establishes the deterministic sequence of results that the
pseudorandom number generator produces.  For instance, contrast the program

```
seed_rng(1234)
for (n in 1:10) print uniform_01_rng()
for (n in 1:10) print uniform_01_rng()
```

which produces the output

```{r}
set.seed(1234)
for (n in 1:10) { cat(rbinom(1, 1, 0.5)); cat(' ') }
for (n in 1:10) { cat(rbinom(1, 1, 0.5)); cat(' ') }
```

with the program

```
seed_rng(1234)
for (n in 1:10) print uniform_01_rng()
seed_rng(1234)
for (n in 1:10) print uniform_01_rng()
```

which produces

```{r}
set.seed(1234)
for (n in 1:10) { cat(rbinom(1, 1, 0.5)); cat(' ') }
set.seed(1234)
for (n in 1:10) { cat(rbinom(1, 1, 0.5)); cat(' ') }
```

Resetting the seed in the second case causes exactly the same ten
pseudorandom numbers to be generated a second time.  Every
well-written pseudorandom number generator and piece of simulation
code should allow the seed to be set manually to ensure reproducibility
of results.^[Replicability of results with different seeds is a
desirable, but stricter condition.]

## Using simulation to estimate event probabilities

We know that $\mbox{Pr}[Y = 1]$ is 0.5 because it represents the flip
of a fair coin.  Simulation based methods allow us to estimate event
probabilities straightforwardly if we can generate random realizations
of the random variables involved in the event definitions.

For example, we know we can generate multiple simulations of flipping
the same coin.  That is, we're not simulating the result of flipping
the same coin ten different times, but simulating ten different
realizations of exactly the same random variable, which represents a
single coin flip.

The fundamental method of computing event probabilities will not
change as we move through this book.  We simply simulate a sequence of
values and return the proportion in which the event occurs as our
estimate.

For example, let's simulate 10 values of $Y$ again and record the
proportion of the simulated values that are 1.  That is, we count the
number of time the event occurs in that the simulated value $y^{(m)}$
is equal to 1.

```
occur = 0
for (m in 1:M)
  y[m] = uniform_01_rng()
  occur = occur + (y[m] == 1)
estimate = occur / M
print `estimated Pr[Y = 1] = ' estimate
```

The equality operator is written as `==`, as in the condition `y[m] ==
1` to distinguish it from the assignment statement `y[m] = 1`, which
sets the value of `y[m]` to 1. The condition expression `y[m] == 1`
evaluates to 1 if the condition is true and 0 otherwise.

If we let `uniform_01_rng(M)` be the result of generating `M`
pseudorandom coin flip results, the program can be shortened to

```
y <- uniform_01_rng(M)
occur = sum(y == 1)
estimate = occur / M
```

A condition such as `y == 1` on a sequence returns a sequence of the
same length with value 1 in positions where the condition is true. For
instance, if

```
y = (2, 1, 4, 2, 2, 1)
```
then
```
y == 2
```
evaluates to
```
(1, 0, 0, 1, 1, 0).
```

Thus `sum(y == 1)` is the number of positions in the sequence `y`
which have the value 1.  Running the program provides the following
estimate based on ten simulation draws.

```{r}
M <- 10
y_sim <- rbinom(M, 1, 0.5);
for (n in 1:M) { cat(y_sim[n]); cat(' '); }
cat('estimated Pr[Y = 1] = ', sum(y_sim) / M)
```
Let's try that a few more times.

```{r}
for (k in 1:10) {
  y_sim <- rbinom(M, 1, 0.5);
  for (n in 1:M) { cat(y_sim[n]); cat(' '); }
  cat(' ')
  cat('estimated Pr[Y = 1] = ', sum(y_sim) / M, '\n')
}
```

The estimates are close, but not very exact.  What if we use 100
simulations?

```{r}
M <- 100
y_sim <- rbinom(M, 1, 0.5);
for (n in 1:M) { cat(y_sim[n]); cat(' '); }
cat('estimated Pr[Y = 1] = ', sum(y_sim) / M)
```

That's closer than most of the estimates based on ten simulation draws.  Let's
try that a few more times without bothering to print all 100 simulated
values,

```{r}
for (k in 1:10) {
  M <- 100
  y_sim <- rbinom(M, 1, 0.5);
  cat('estimated Pr[Y = 1] = ', sum(y_sim) / M, '\n')
}
```

What happens if we let $M = 10,000$ simulations?

```{r}
for (k in 1:10) {
  M <- 10000
  y_sim <- rbinom(M, 1, 0.5);
  cat('estimated Pr[Y = 1] = ', sum(y_sim) / M, '\n')
}
```

Now the estimates are very close to the true probability being
estimated (i.e., 0.5, because the flip is fair). This raises the
questions of how many simulation draws we need in order to be
confident our estimates are close to the values being estimated.






## Law of large numbers

Visualization in the form of simple plots goes a long way toward
understanding concepts in statistics and probability.  A traditional
way to plot what happens as the number of simulation draws $M$
increases is to keep a running tally of the estimate as each draw is
made and plot the estimated event probability $\mbox{Pr}[Y = 1]$ for
each $m \in 1:M$.^[See, for example, the quite wonderful little book,
Bulmer, M.G., 1965. *Principles of Statistics*. Oliver and Boyd,
Edinburgh.]

To calculate such a running tally of the estimate at each online, we can do this:

```
occur = 0
for (m in 1:M)
  y[m] = uniform_01_rng(M)
  occur = occur + (y[m] == 1)
  estimate[m] = occur / m
```

Recall that the expression `(y[m] == 1)` evaluates to 1 if the
condition holds and 0 otherwise. The result of running the program is
that `estimate[m]` will hold the estimate $\mbox{Pr}[Y = 1]$ after $m$
simulations.  We can then plot the estimates as a function
of the number of draws using a line plot to display the trend.

```{r}
set.seed(0)
library(ggplot2)
M <- 1e5
Ms <- c()
y_sim <- rbinom(M, 1, 0.5)
hat_E_Y <- c()
Ms <- c()
for (i in 0:50) {
  Ms[i + 1] <- min(M, (10^(1/10))^i)
  hat_E_Y[i + 1] <- mean(y_sim[0:Ms[i + 1]])
}
df <- data.frame(M = Ms, hat_E_Y)
```

```{r fig.cap="Monte Carlo estimate of probability that a coin lands head as a function of the number of simulation draws.  The line at 0.5 marks the true probability being estimated.  Plotted on a linear scale, it is clear how quickly the estimates converge to roughly the right value."}
plot <- ggplot(df, aes(x = M, y = hat_E_Y)) +
  geom_hline(yintercept = 0.5, color = "red") +
  geom_line() +
  geom_point() +
  scale_x_continuous(breaks=c(1, 50000, 100000), limits=c(1, 100000),
                     labels=c("1", "50,000", "100,000")) +
  scale_y_continuous(limits = c(0, 1),
                     breaks = c(0, 0.25, 0.5, 0.75, 1.0)) +
  xlab("simulation draws") +
  ylab("estimated Pr[Y = 1]") +
  ggtheme_tufte()
plot
```

The linear scale of the previous plot obscures the behavior of the estimates.  Consider instead a plot with the $x$-axis on the logarithmic scale.

```{r fig.cap="Monte Carlo estimate of probability that a coin lands head as a function of the number of simulation draws.  The line at 0.5 marks the true probability being estimated.  The log-scaled $x$-axis makes the early rate of convergence more evident."}
plot <- ggplot(df, aes(x = M, y = hat_E_Y)) +
  geom_hline(yintercept = 0.5, color = "red") +
  geom_line() +
  geom_point() +
  scale_x_log10(breaks=10^(0:5), limits=c(1, 100000),
                labels=c("1", "10", "100", "1,000", "10,000", "100,000")) +
  scale_y_continuous(limits = c(0, 1),
                     breaks = c(0, 0.25, 0.5, 0.75, 1.0)) +
  xlab("simulation draws") +
  ylab("estimated Pr[Y = 1]") +
  ggtheme_tufte()
plot
```

With a log-scaled $x$-axis, the values between 1 and 10 are plotted
with the same width as the values between $10\,000$ and $100\,000$;
both take up 20% of the width of the plot. On the linear scale, the
values between 1 and 10 take up only $\frac{10}{100\,000}$, or 0.01%$,
of the plot, wehreas values between $10\,000$ and $100\,000$ take up
90% of the plot.

Plotting the progression of multiple simulations demonstrates the
trend in errors.

```{r fig.cap="Each of the one hundred grey lines represents the ratio of heads observed in a sequence of coin flips, the size of which indicated on the $x$-axis.  The line at 0.5 indicates the probability a coin lands heads in a fair coin toss.  The convergence of the ratio of heads to 0.5 in all of the sequences is clearly visible."}

library(ggplot2)
set.seed(0)
M_max <- 1e4
J <- 100
I <- 47
N <- I * J
df2 <- data.frame(r = rep(NA, N), M = rep(NA, N), hat_E_Y = rep(NA, N))
pos <- 1
for (j in 1:J) {
  y_sim <- rbinom(M_max, 1, 0.5)
  for (i in 4:50) {
    M <- max(100, min(M_max, (10^(1/10))^i))
    hat_E_Y = mean(y_sim[1:M])
    df2[pos, ] <- list(r = j, M = M, hat_E_Y = hat_E_Y);
    pos <- pos + 1;
  }
}

pr_Y_eq_1_plot <- ggplot(df2, aes(x = M, y = hat_E_Y, group=r)) +
    geom_hline(yintercept = 0.5, color = "red") +
    geom_line(alpha=0.15) +
    scale_x_log10(breaks=10^(2:4), limits=c(100, 10000),
                  labels=c("100", "1,000", "10,000")) +
    scale_y_continuous(limits = c(0.375, 0.625),
                       breaks = c(0.4, 0.5, 0.6),
                       labels = c(0.4, 0.5, 0.6)) +
    xlab("simulation draws") +
    ylab("estimated Pr[Y = 1]") +
    ggtheme_tufte()
pr_Y_eq_1_plot
```

```{r fig.cap="Continuing where the previous plot left off, each of the one hundred grey lines represents the ratio of heads observed in a sequence of coin flips.  The values on the $x$ axis is one hundred times larger than in the previous plot, and the scale of the $y$-axis is one tenth as large.  The trend in error reduction appears the same at the larger scale."}

library(ggplot2)
set.seed(0)
M_max <- 1e6
J <- 100
I <- 47
N <- I * J
df2 <- data.frame(r = rep(NA, N), M = rep(NA, N), hat_E_Y = rep(NA, N))
pos <- 1
for (j in 1:J) {
  y_sim <- rbinom(M_max, 1, 0.5)
  for (i in 4:60) {
    M <- max(100, min(M_max, (10^(1/10))^i))
    hat_E_Y = mean(y_sim[1:M])
    df2[pos, ] <- list(r = j, M = M, hat_E_Y = hat_E_Y);
    pos <- pos + 1;
  }
}

pr_Y_eq_1_plot <- ggplot(df2, aes(x = M, y = hat_E_Y, group=r)) +
    geom_hline(yintercept = 0.5, color = "red") +
    geom_line(alpha=0.15) +
    scale_x_log10(breaks=10^(4:6), limits=c(1e4, 1e6),
                  labels=c("10,000", "100,000", "1,000,000")) +
    scale_y_continuous(limits = c(0.485, 0.515),
                       breaks = c(0.49, 0.5, 0.51),
                       labels = c(0.49, 0.5, 0.51)) +
    xlab("simulation draws") +
    ylab("estimated Pr[Y = 1]") +
    ggtheme_tufte()
pr_Y_eq_1_plot
```

The *law of large numbers*^[Which technically comes in a strong and
weak form.] says roughly that as the number of simulated values grows,
the average will converge to the expected value.  In this case, our
estimate of $\mbox{Pr}[Y = 1]$ can be seen to converge to the true
value of 0.5 as the number of simulations $M$ increases.
Because the quantities involved are probabilistic, the exact specification is a
little more subtle than the $\epsilon$-$\delta$ proofs in calculus.





## Simulation notation

We will use parenthesized superscripts to pick out the elements of a sequence of simulations.  For example,

$$
y^{(1)}, y^{(2)}, \ldots, y^{(M)}
$$

will be used for $M$ simulations of a single random variable
$Y$.^[Each $y^{(m)}$ is a possible realization of $Y$, which is why
they are written using lowercase.]  It's important to keep in mind
that this is $M$ simulations of a single random variable, not a single
simulation of $M$ different random variables.

Before we get going, we'll need to introduce *indicator function*
notation.  For example, we write

$$
\mathrm{I}[y^{(m)} = 1]
=
\begin{cases}
1 & \mbox{if} \ y^{(m)} = 1
\\[4pt]
0 & \mbox{otherwise}
\end{cases}
$$

The indicator function maps a condition, such as $y^{(m)} = 1$ into
the value 1 if the condition is true and 0 if it is false.^[Square
bracket notation is used for functions when the argument is itself a
function.  For example, we write $\mbox{Pr}[Y > 0]$ because $Y$ is a
random variable, which is modeled as a function.  We also write
$\mathrm{I}[x^2 + y^2 = 1]$ because the standard bound variables $x$
and $y$ are functions from contexts defining variable values.]

Now we can write out the formula for our estimate of $\mbox{Pr}[Y =
1]$ after $M$ draws,

$$
\begin{array}{rcl}
\mbox{Pr}[Y = 1]
& \approx &
\frac{\displaystyle \mathrm{I}[y^{(1)} = 1]
      \ + \ \mathrm{I}[y^{(2)} = 1]
      \ + \ \cdots
      \ + \ \mathrm{I}[y^{(M)} = 1]}
     {\displaystyle M}
\end{array}
$$

That is, our estimate is the proportion of the simulated values which
take on the value 1.  It quickly becomes tedious to write out
sequences, so we will use standard summation notation, where we write

$$
\mathrm{I}\!\left[y^{(1)} = 1]
      + \mathrm{I}[y^{(2)} = 1]
      + \cdots
      + \mathrm{I}[y^{(M)} = 1\right]
\ = \
\sum_{m=1}^M \mathrm{I}[y^{(m)} = 1]
$$

Thus we can write our simulation-based estimate of the probability
that a fair coin flip lands heads as^[In general, the way to estimate an event probability $\phi(Y)$ where $\phi$ defines some condition, given simulations $y^{(1)}, \ldots, y^{(M)}$ of $Y$, is as $$\mbox{Pr}[\phi(Y)] = \frac{1}{M} \sum_{m = 1}^M \mathrm{I}[\phi(y^{(m)})].$$]

$$
\mbox{Pr}[Y = 1]
\approx
\frac{1}{M}
\,
\sum_{m=1}^M \mathrm{I}[y^{(m)} = 1]
$$

The form $\frac{1}{M} \sum_{m=1}^M$ will recur repeatedly in
simulation --- it just says to average over values indexed by $m \in
1:M$.^[We are finally in a position to state the *strong law of large numbers* as the event probability of a limit, $$\mbox{Pr}\!\left\lbrack \lim_{M \rightarrow \infty} \frac{1}{M} \sum_{m = 1}^M \mathrm{I}\!\left\lbrack y^{(m)} = 1 \right\rbrack \ = \ 0.5 \right\rbrack,$$  where each $y^{(m)}$ is a separate fair coin toss. ]


## Central limit theorem

The law of large numbers tells us that with more simulations, our
estimates become more and more accurate.  But they do not tell us how
quickly we can expect that convergence to proceed.  The *central limit
theorem* provides the convergence rate.

First, we have to be careful about what we're defining.  First, we
define the *error* for an estimate as the difference from the true
value,

$$
\left( \frac{1}{M} \sum_{m=1} \mathrm{I}[y^{(m)} = 1] \right) - 0.5
$$

The *absolute error* is just the absolute value^[In general, the
absolute value function applied to a real number $x$ is written as
$|x|$ and defined to be $x$ if $x$ is non-negative and $-x$ if $x$ is
negative.] of this,

$$
\left| \,
\left( \frac{1}{M} \sum_{m=1} \mathrm{I}[y^{(m)} = 1] \right) - 0.5
\, \right|
$$


```{r fig.cap="The absolute error of the simulation-based probability estimate as a function of the number of simulation draws.  One hundred sequences of one million flips are shown."}

set.seed(1234)
M_max <- 1e6
J <- 100
I <- 47
N <- I * J
df2 <- data.frame(r = rep(NA, N), M = rep(NA, N), err_hat_E_Y = rep(NA, N))
pos <- 1
for (j in 1:J) {
  y_sim <- rbinom(M_max, 1, 0.5)
  for (i in 4:60) {
    M <- max(100, min(M_max, (10^(1/10))^i))
    err_hat_E_Y = abs(mean(y_sim[1:M]) - 0.5)
    df2[pos, ] <- list(r = j, M = M, err_hat_E_Y = err_hat_E_Y);
    pos <- pos + 1;
  }
}

abs_err_plot <-
  ggplot(df2, aes(x = M, y = err_hat_E_Y, group = r)) +
  geom_hline(yintercept = 0.5, color = "red") +
  geom_line(alpha=0.15) +
  scale_x_log10(breaks=10^(4:6), limits=c(1e4, 1e6),
                labels=c("10,000", "100,000", "1,000,000")) +
  scale_y_continuous(limits = c(0, 0.015),
                     breaks = c(0, 0.05, 0.01),
                     labels = c("0.00", "0.05", "0.01")) +
  xlab("simulation draws") +
  ylab("absolute error") +
  ggtheme_tufte()
abs_err_plot
```



```{r}
set.seed(1234)
M_max <- 1e6
J <- 300
Ms <- 10^((6:12) / 2)
N <- length(Ms)
ys <- matrix(NA, N, J)
for (j in 1:J) {
  z <- rbinom(M_max, 1, 0.5)
  for (n in 1:N) {
    ys[n, j] <- abs(mean(z[1:Ms[n]]) - 0.5)
  }
}
mean <- rep(NA, N)
sd <- rep(NA, N)
quant_68 <- rep(NA, N)
quant_95 <- rep(NA, N)
sd <- rep(NA, N)
sixty_eight_pct = pnorm(1) - pnorm(-1)
ninety_five_pct = pnorm(2) - pnorm(-2)
for (n in 1:N) {
  mean[n] <- mean(ys[n, ])
  sd[n] <- sd(ys[n, ])
  quant_68[n] <- quantile(ys[n, ], sixty_eight_pct)
  quant_95[n] <- quantile(ys[n, ], ninety_five_pct)
}
```

Plotting both the number of simulations and the absolute error on the
log scale reveals the rate at which the error decreases with more
draws.

```{r fig.cap="Absolute error versus number of simulation draws for 100 simulated sequences of $M = 1\\,000\\,000$ draws. The blue line is at the 68 percent quantile and the red line at the 95 percent quantile of these draws.  The relationship between the log number of draws and log error is revealed to be linear."}
#   commented out bits plot the clt version
#   geom_line(data = data.frame(x = c(1e4, 1e6),
#                               y = 2 * c(.5 / sqrt(1e4), .5 / sqrt(1e6))),
#             aes(x = x, y = y, group = NA),
#             color = "red") +
#   geom_line(data = data.frame(x = c(1e4, 1e6),
#                               y = c(.5 / sqrt(1e4), .5 / sqrt(1e6))),
#             aes(x = x, y = y, group = NA),
#             color = "blue") +

fudge <- 1e-6
x <- 0.5 * (1:100)
log_abs_err_plot <- ggplot(df2, aes(x = M, y = err_hat_E_Y + fudge, group=r)) +
  geom_line(alpha=0.15) +
  geom_line(data = data.frame(Ms = Ms, quantile = quant_68),
            aes(x = Ms, y = quantile, group = NA),
            color = "blue", size = 0.5) +
  geom_line(data = data.frame(Ms = Ms, quantile = quant_95),
            aes(x = Ms, y = quantile, group = NA),
            color = "red", size = 0.5) +
  scale_x_log10(limits=c(.9 * 1e3, 1e6 * 1.1),
                breaks=10^(3:6),
                labels=c("1,000", "10,000", "100,000", "1,000,000")) +
  scale_y_log10(limits = c(0.8e-4, 2e-1),
                breaks = c(0.0001, 0.001, 0.01, 0.1),
                labels = c("0.0001", "0.001", "0.01", "0.1" )) +
  xlab("simulation draws") +
       ylab("log absolute error + 1e-5") +
  ggtheme_tufte()
log_abs_err_plot
```

We can read two points $(x_1, y_1)$ and $(x_2, y_2)$ off of the graph for $x_1 = 10\,000$ and $x_2 = 100\,000$ as

```{r}
printf("x[1], y[1] = %7.f, %6.5f\nx[2], y[2] = %7.f, %6.5f",
       Ms[3], quant_68[3],
       Ms[5], quant_68[5])
```

which gives us the following values on the log scale

```{r}
printf("log x[1], log y[1] = %5.2f, %4.2f\nlog x[2], log y[2] = %5.2f, %4.2f",
       log(Ms[3]), log(quant_68[3]),
       log(Ms[5]), log(quant_68[5]))
```

Using the log scale values, the estimated slope of the reduction in quantile bounds is

```{r}
printf("estimated slope\n(log y[2] - log y[1])\n / (log x[2] - log x[1])  =  %3.2f",
       (log(quant_68[5]) - log(quant_68[3])) / (log(Ms[5]) - log(Ms[3])) )
```

If we let $\epsilon_M$ be the value of one of the quantile lines at
$M$ simulation draws, the linear relationship plotted in the figures
have the form^[A line through the points $(x_1, y_1)$ and $(x_2, y_2)$ has
$$
\mbox{slope} = \frac{y_2 - y_1}{x_2 - x_1}.
$$]

$$
\log \epsilon_M = -\frac{1}{2} \, \log M + \mbox{const}.
$$

When writing "const" in a mathematical expression, the presumption is
that it refers to a constant that does not depend on the free
variables of interest (here, $M$, the number of simulation draws).
Ignoring constants lets us focus on the order of the dependency.  The
red line and blue line have the same slope, but different constants.

Seeing how this plays out on the linear scale requires exponentiating
both sides of the equation and reducing,

$$
\begin{array}{rcl}
\exp(\log \epsilon_M)
& = &
\exp( -\frac{1}{2} \, \log M + \mbox{const} )
\\[6pt]
%
\epsilon_M
& = &
\exp( -\frac{1}{2} \, \log M )
\times
\exp(\mbox{const})
%
\\[6pt]
\epsilon_M
& = &
\exp( \log M )^{-\frac{1}{2}}
\times
\exp(\mbox{const})
\\[6pt]
\epsilon_M
& = &
\exp(\mbox{const}) \times M^{-\frac{1}{2}}
\end{array}
$$

Dropping the constant, this relationship between the expected absolute
error $\epsilon_M$ after $M$ simulation draws may be succinctly
summarized using proportionality notation,^[In general, we write $$f(x)
\propto g(x)$$ if there is a positive constant $c$ that does not depend
on $x$ such that $$f(x) = c \times g(x).$$ For example, $$3x^2 \propto
9x^2,$$ with $c = \frac{1}{3}$.]

$$
\displaystyle
\epsilon_M
\ \propto \
\frac{\displaystyle 1}{\displaystyle \sqrt{M}}.
$$

This is a fundamental result in statistics derived from the
*central limit theorem*. The central limit theorem governs the
accuracy of almost all simulation-based estimates. We will return to a
proper formulation when we have the scaffolding in place to deal with
the pesky constant term.

In practice, what does this mean? It means that if we want to get an
extra decimal place of accuracy in our estimates, we need one hundred
(100) times as many draws. For example, the plot shows error rates
bounded by roughly 0.01 with $10\,000$ draws, yielding estimates that
are very likely to be within $(0.49, 0.51).$ To reduce that likely
error bound to 0.001, that is, ensuring estimates are very likely in
$(0.0499, 0.501),$ requires 100 times as many draws (i.e., a whopping
$1\,000\,000$ draws).^[For some perspective, $10\,000$ is the number
of at bats in an entire twenty-year career for a baseball player,
the number of field goal attempts in an entire career of most basketball
players, and the size of a very large disease study or meta-analysis in
epidemiology.]

Usually, one hundred draws will provide a good enough estimate of most
quantities of interest in applied statistics. The variability of the
estimate based on a single draw depends on the variability of the
quantity being estimated. One hundred draws reduces the expected
estimation bound to one tenth of the variability of a single draw.
Reducing that variability to one hundredth of the variability of a
single draw would require ten thousand draws. In most applications,
the extra estimation accuracy is not worth the extra computation.
