# Rejection Sampling

## Inverse cumulative distribution generation


We have so far assumed we have a uniform random number generator that
can sample from a $\mbox{uniform}(0, 1)$ distribution. That
immediately lets us simulate from a $\mbox{uniform}(a, b).$^[If $$ U
\sim \mbox{uniform}(0, 1) $$ then $$a + U \times (b - a) \sim
\mbox{uniform}(a, b).$$] But what if we want to simulate realizations
of a random variable $Y$ whose density $p_Y$ is not uniform?

If we happen to have a random variable $Y$ for which we can compute
the inverse $F^{-1}_Y(p)$ of the cumulative distribution function for
$p \in (0, 1),$^[If $F^{-1}_Y(p) = y$ then $F^{-1}(y) =
\mbox{Pr}[Y \leq y] = p.$] then we can simulate random realizations of
$Y$ as follows.  First, simulate a uniform draw $U$,

$$
u^{(m)} \sim \mbox{uniform}(0, 1),
$$

then apply the inverse cumulative distribution function to turn it
into a simulation of $Y$,

$$
y^{(m)} = F^{-1}(u^{(m)}).
$$

It turns out we've already seen an example of this strategy---it is
how we we simulated from a logistic distribution right off the bat.
The log odds transform is the inverse cumulative distribution function
for the logistic distribution. That is, if

$$
Y \sim \mbox{logistic}(0, 1),
$$

then

$$
F_Y(y) = \mbox{logit}^{-1}(y) = p
$$

and hence

$$
F^{-1}_Y(p) = \mbox{logit}(p) = y.
$$

This lets us simulate $Y$ by first simulating $U \sim
\mbox{uniform}(0, 1)$, then setting $Y = \mbox{logit}(U)$. Both the
log odds function and its inverse are easy to compute. Often, the
inverse cumulative distribution function is an expensive
computation.^[In general, the inverse cumulative distribution function
$F^{-1}_Y(p)$ can be computed by solving $$F_Y(u) = \int_{\infty}^u
p_Y(y) \, \mathrm{d} y = p$$ for $u$, which can typically be
accomplished numerically if not analytically.]


## Beta distribution

As a first non-trivial example, let's consider the beta distribution,
which arises naturally as a posterior in binary models. Recall that
when we have a binomial likelihood $\mbox{binomial}(y \mid N, \theta)$
and a uniform prior $\theta \sim \mbox{uniform}(0, 1)$, then the
posterior density is

$$
p(\theta \mid y)
\ \propto \
\theta^y \times (1 - \theta)^{N - y}.
$$

This turns out to be a beta distribution with parameters $\alpha = y +
1$ and $\beta = N - y + 1$. If a random variable $\Theta
\in (0, 1)$ has a *beta distribution*, then its density $p_{\Theta}$ is

$$
\mbox{beta}(\theta \mid \alpha, \beta)
\ \propto \
\theta^{\alpha - 1} \times (1 - \theta)^{\beta - 1}
$$

for some $\alpha, \beta > 0$.^[We prefer to present densities up to
normalizing constants, because the normalizing constants are
distracting---what matters is how the density changes with the
variate. Here, the fully normalized distribution is
$$\mbox{beta}(\theta \mid \alpha, \beta) \ = \
\frac{1}{\mbox{B}(\alpha, \beta)} \, \theta^{\alpha - 1} \times (1 -
\theta)^{\beta - 1},$$ where Euler's beta function, which gives its
namesake distribution its name, is defined as the integral of the
unnormalized beta density $$ \mathrm{B}(\alpha,
\beta) \ = \ \displaystyle \int_0^1 \theta^{\alpha - 1} (1 -
\theta)^{\beta - 1} \, \mbox{d} \theta.$$  Hence, it's clear the
beta density integrates to 1 for any $\alpha, \beta$.]

If $Y \sim \mbox{beta}(\alpha, \beta)$, then its expected value is

$$
\mathbb{E}[Y] = \frac{\alpha}{\alpha + \beta}.
$$

It is convenient to work with the beta distribution parameters in
terms of the mean $\alpha / (\alpha + \beta)$ and the total count
$\alpha + \beta$.^[Distributions have means, random variables have
expectations.  The mean of a distribution is the expectation of a
random variable with that distribution, so the terms are often
conflated.] The higher the total count, lower the variance. For
example, here is a plot of a few beta distributions organized by total
count and mean.

```{r, out.width = "100%", fig.width=9, fig.asp = 0.8, fig.cap="Plots of the densities $\\mbox{beta}(\\theta \\mid \\alpha, \\beta)$ for different total counts $\\alpha + \\beta$ and different means $\\alpha / (\\alpha + \\beta).$ The vertical dotted lines are drawn at the mean. Only the distribution with 0.5 mean has a symmetric density; the other densities are skewed toward the right and thus have means to the left of their modes (maxima), where the modes exist."}

beta_density_df <-
  data.frame(theta = c(), p_theta = c(), count = c(), mean = c())
mean_df <- data.frame(expectation = c(), count = c(), mean = c())
thetas = (0:100)/100
for (kappa in c(0.5, 2, 8, 32)) {
  for (mu in c(0.5, 0.675, 0.85)) {
    df_temp <-
          data.frame(theta = thetas,
                     p_theta = dbeta(thetas, kappa * mu, kappa * (1 - mu)),
                     count = rep(paste("count ", kappa), length(thetas)),
		     mean = rep(paste("mean ", mu), length(thetas)),
		     mu = rep(mu, length(thetas)))
    beta_density_df <- rbind(beta_density_df, df_temp)
  }
}
beta_density_plot <-
  ggplot(beta_density_df, aes(x = theta, y = p_theta)) +
  facet_grid(mean ~ count) +
  geom_line() +
  geom_hline(yintercept = 0, linetype = "dashed", size = 0.25) +
  geom_vline(aes(xintercept = mu), linetype = "dotted") +
  scale_x_continuous(lim = c(0, 1), breaks = c(0, 0.5, 1),
                     labels = c("0", "0.5", "1")) +
  scale_y_continuous(lim = c(0, 7), breaks = c(0, 3, 6)) +
  ggtheme_tufte() +
  xlab(expression(theta)) +
  ylab(expression(
    paste("beta(", theta, " | ", alpha, ", ", beta, ")"))) +
  theme(panel.spacing.x = unit(3, "lines"),
        panel.spacing.y = unit(2, "lines"))
beta_density_plot
```

For example, a total count of $\alpha + \beta = 8$ and mean of $\alpha
/ (\alpha + \beta) = 0.85$ corresponds to beta distribution parameters
$\alpha = 8 \times 0.85 = 6.8$ and $\beta = 8 \times (1 - 0.85) =
1.2.$

The plot shows that when the mean is 0.5 and the count is 2 (i.e.,
$\alpha = \beta = 1$), the result is a uniform distribution.  The
algebra agrees,

$$
\begin{array}{rcl}
\mbox{beta}(\theta \mid 1, 1)
& \propto & \theta^{1 - 1} \times (1 - \theta)^{1 - 1}
\\[4pt]
& = & 1
\\[4pt]
& = & \mbox{uniform}(\theta \mid 0, 1).
\end{array}
$$

The area under each of these curves, as drawn, is exactly one. The
beta distributions with small total count ($\alpha + \beta$)
concentrate most of their probability mass near the boundaries. As the
count grows, the probability mass concentrates away from the
boundaries and around the mean.^[Nevertheless, in higher dimensions,
the curse of dimensionality rears its ugly head, and concentrates the
total mass in the corners, even if each dimension is independently
distributed and drawn from a distribution concentrated away from the
edges in one dimension. The reason is the same---each dimension's
value squared just pulls the expected distance of a draw further from
the multidimensional mean.]

Each of these distributions has a well-defined mean, as shown in row
labels in the plot. But they do not all have well defined modes
(maxima). For example, consider the beta distribution whose density is
shown in the upper left example in the plot. It shows a U-shaped
density for a $\mbox{beta}(0.25, 0.25)$ distribution, which
corresponds to mean $0.5 = 0.25 / (0.25 + 0.25)$ and total count $0.5
= 0.25 + 0.25$. As $\theta$ approaches either boundary, 0 or 1, the
density grows without bound. There is simply no maximum value for the
density.^[Despite the lack of a maximum, the area under the density is
one. The region of very high density near the boundary becomes
vanishingly narrow in order to keep the total area at one.]

## Uniform, bounded rejection sampling

How do we sample from a beta distribution?  Rejection sampling is
very simple algorithm to sample from general distributions for which
we know how to compute the density.  It is a good starter algorithm
because it is easy to understand, but points the way toward more
complex sampling algorithms we will consider later.

Let's start with the simplest possible kind of rejection sampling
where we have a bounded distribution like a beta distribution. The
values drawn from a beta distribution are bounded between 0 and 1 by
construction. Furthermore, if $\alpha, \beta > 1$, as we will assume
for the time being, then there is a maximum value for
$\mbox{beta}(\theta \mid \alpha, \beta)$.^[The maximum density occurs
at the *mode* of the distribution, which for
$\\mbox{beta}(\\theta \\mid \\alpha, \\beta)$ is given by
$$\\theta^* = \\frac{\\alpha - 1}{\\alpha + \\beta - 2}.$$]


For concreteness, let's start specifically with a $\mbox{beta}(6.8,
1.2)$ distribution, corresponding to the mean 0.85 and count of 8 case
from the previous section. We observe that all values fall below 5, so
we will create a box with height 5 and width of 1.^[For an unknown
distribution, we could use an optimization algorithm to find the
largest value.] Next, we draw uniformly from the box, drawing a
horizontal position $\theta^{(m)} \sim \mbox{uniform}(0, 1)$ and
vertical position $u^{(m)} \sim \mbox{uniform}(0, 5)$. The points
whose value for $u$ falls below the density at the value for $\theta$
are retained.

```{r fig.cap = 'A simple instance of rejection sampling from a bounded $\\mbox{beta}(6.8, 1.2)$ distribution, whose density is shown as a solid line.  Points $(\\theta, u)$ are drawn uniformly from the rectangle, then accepted as a draw of $\\theta$ if $u$ falls below the density at $\\theta$.  The accepted draws are rendered as plus signs and the rejected ones as circles.  The acceptance rate here is roughly 20 percent.'}

M <- 500
alpha <- 6.8
beta <- 1.2
mean = alpha / (alpha + beta)
mode = (alpha - 1) / (alpha + beta - 2)
y <- (0:100) / 100
u <- runif(M, 0, 5)
theta <- runif(M, 0, 1)
accept <- (u < dbeta(theta, alpha, beta))

reject_beta_df <-
  data.frame(y = y,
             p_y = dbeta(y, alpha, beta))
accept_beta_df <-
  data.frame(theta = theta, u = u, accept = accept)

reject_beta_plot <-
  ggplot(accept_beta_df) +
  geom_line(aes(x = y, y = p_y), data = reject_beta_df) +
  geom_point(aes(x = theta, y = u, shape = accept),
             color = "#333333", size = 1.5) +
  scale_shape_manual(values = c(1, 3)) +
  geom_vline(xintercept = c(0, 1), linetype="dotted") +
  geom_hline(yintercept = 5, linetype = "dotted") +
  geom_hline(yintercept = 0, linetype = "dotted") +
  xlab(expression(theta)) +
  ylab(expression(u)) +
  ggtheme_tufte() +
  theme(legend.position = "none")
reject_beta_plot
```

More specifically, we keep the values $\theta^{(m)}$ where

$$
u^{(m)} < \mbox{beta}(\theta^{(m)} \mid 6.8, 1.2).
$$

The set of accepted draws that are retained are distributed uniformly
in the area under the density plot. The probability of a draw from rejection
sampling for variable $U$ falling between points $a$ and $b$ is
proportional to the area under the density curve between $a$ and $b$,
which is the correct probability by the definition of the probability
density function $p_U$, which ensures

$$
\mbox{Pr}[a \leq U \leq b] = \int_a^b p_U(u) \, \mathrm{d}u.
$$

If the intervals are right, the density is right.


## Rejection sampling algorithm

To generate a single draw from $\mbox{beta}(6.8, 1.2)$, we continually
sample points $(u, \theta)$ uniformly until we find one where $u$
falls below the density value for $\theta$, then return the $\theta$
value. Because the variable $u$ is only of use for sampling, it is
called an *auxiliary variable*. Many sampling methods use auxiliary
variables. The rejection sampling algorithm written out for our
particular case is as follows.

```
while (true)
  u = uniform_rng(0, 5)
  theta = uniform_rng(0, 1)
  if (u < beta(theta | 6.8, 1.2))
    return theta
```

Let's run this algorithm for $M = 100\,000$ iterations and see what the
histogram looks like.

```{r, fig.cap="Histogram of $M = 100\\,000$ Draws from $\\mbox{beta}(6.8, 1.2)$ made via rejection sampling.  The true density is plotted over the histogram as a line.  The acceptance rate for draws was roughly 20 percent."}

set.seed(1234)
M <- 100000
accept <- 0
total <- 0
m <- 1
theta <- rep(NA, M)
while (m <= M) {
  u <- runif(1, 0, 5)
  theta_star <- runif(1, 0, 1)
  total <- total + 1
  if (u < dbeta(theta_star, 6.8, 1.2)) {
    accept <- accept + 1
    theta[m] <- theta_star
    m <- m + 1
  }
}
beta_draws_df <- data.frame(theta = theta)
beta_draws_plot <-
  ggplot(beta_draws_df, aes(x = theta)) +
  geom_histogram(stat = "density", n = 80, color = "black", fill = "#ffffe8",
                 size = 0.15) +
  stat_function(fun = dbeta, args = list(shape1 = 6.8, shape2 = 1.2),
               size = 0.35, color = "black") +
  xlab(expression(theta)) +
  ylab(expression(p[Theta](theta))) +
  ggtheme_tufte()
beta_draws_plot
# printf("acceptance percentage = %3.2f\n", accept/total)
```

This looks like it's making the appropriately distributed draws from
the beta distribution.^[After we introduce the normal distribution, we
will develop a $\chi^2$ test statistic for whether a given set of
draws come from a specified distribution. For now, we perform the
inaccurate test known informally as "$\chi$ by eye."]

## Acceptance, concentration, and the curse of dimensionality

As noted in the caption of the plot, the acceptance rate is only 20%
for the uniform proposals. This acceptance rate can become arbitrarily
bad with uniform proposals as the true distribution from which we want
to sample concentrates around a single value. As such, rejection
sampling's value is as a pedagogical example for introducing generic
sampling and also as a component in more robust sampling algorithms.

The algorithm could also be extended to higher dimensions, but the
problem of rejection becomes worse and worse due to the curse of
dimensionality. If the posterior is concentrated into a reasonably
small area (even as mildly in our $\mbox{beta}(6.8, 1.2)$ example) in
each dimension, the chance of a random multidimensional sample being
accepted in every dimension is only $0.2^N$, which becomes vanishingly
small for practical purposes even in 10 dimensions.^[In ten
dimensions, overall acceptance of a uniform draw would be $0.2^{10}$, or
a little less than one in ten million.  While this may be possible
with patience, another ten dimensions becomes completely unworkable,
even with massive parallelism.]


## General rejection sampling

The more general form of rejection sampling takes a proposal from an
arbitrary scaled density. In the example of the previous section, we
used a $\mbox{uniform}(0, 1)$ distribution scaled by a factor of five.
The acceptance procedure in the general case remains exactly the same.

In the general case, suppose we want to draw from a density $p(y)$.
We'll need a density $q(y)$ from which we know how to simulate draws,
and we'll need a constant $c$ such that

$$
c \times q(y) > p(y)
$$

for all $y$.^[This means the support of the proposal distribution must
be at least as large as that of the target distribution.]

The general rejection sampling algorithm for target density $p(y)$,
proposal density $q(y)$ with constant $c$ such that $c \times q(y) >
p(y)$ for all $y$, is as follows:

```
while (true)
  y = q_rng(y)
  u = uniform(0, c * q(y))
  if (u < p(y)) return y
```

Our simplified algorithm in the previous section was just a special
case where $q(y)$ is uniform over a bounded interval. The argument for
correctness in the more general case is identical---draws are made
proportional to the area under the density, which provides the correct
distribution.