# Pseudorandom Number Generators

We have been assuming so far that we have pseudorandom number
generators that work in some sense to simulate true randomness.  This
chapter clarifies what that means and shows how we gain confidence in
our random number generator's randomness through testing various
hypotheses that it's random against the actual results produced.

## Linear congruential generator

To provide a feel for what a (weak) pseudorandom number generator
looks like, we'll start with a very simple one.  Starting from an
initial random number *seed* $s$, the *linear congruential*
pseudorandom number generator does a linear transform and then reduces
the number modulo $M$ to generate an integer in the range $0, 1,
\ldots, M - 1$.^[The *modulus operator* is defined so that $m \bmod n$
is the remainder after dividing $m$ by $n$.  For example, $5 \bmod 2 =
1$ and $6 \bmod 3 = 0$.]  The sequence of random numbers generated,
$y_0, y_2, \ldots$, is defined inductively based on an integer seed
$m$, integer multiplier $\alpha$, and integer increment $\beta$, by the base case

$$
y_0 = s \bmod M
$$

and inductive case

$$
y_{n + 1} = \left( \alpha \times y_n + \beta \right) \bmod M.
$$

The pseudocode for generating the next random number given the
previous random number is a one-liner.^[Such a function is typically
set up as an iterator with either object-encapsulated or static storge
of its arguments.]

```
next_prng(seed, M, alpha, beta)
  return (alpha * last_y + beta) mod M
```

We will conveniently choose $M = 2^16$ to provide 16-bit integer
results.^[We are implementing these algorithms in R, which has the
serious limitation of restricting user integer values to short, 32-bit
values.]

To make sure this works to rotate through all $M$ possible values, we
need $\beta$ to be relatively prime to $M$.  So we'll just set $a =
123$ and and $b = 127$ and start with a seed of $s = 1234$.

```{r}
next_prng <- function(y, M, a, b) as.integer((a * y + b) %% M)
M <- 2^16
a <- 123
b <- 117
s <- 1234
for (j in 1:4) {
  for (k in 1:6) {
    u <- next_prng(s, M, a, b)
    printf("%10d", u)
    s <- u
  }
  printf("\n")
}
```

Let's go ahead and generate $10\,000$ draws and plot a histogram.^[For
this to look roughly uniform, as here, it's important to set the
boundary of the first bin at 0, set the limit to be exactly $0$
to $2^{16} - 1$, and to make all the bins the same width.  We achieve
the latter by choosing the number of bins to be 16, a multiple of the
total number of possible outcomes, $2^{16} - 1$.]

```{r}
M <- 2^16
a <- 123
b <- 117
s <- 1234
N <- 10000
y <- rep(NA, N)
for (n in 1:N) {
  y[n] <- next_prng(s, M, a, b)
  s <- y[n]
}

lc_prng_df <- data.frame(y = y)
lc_prng_plot <-
  ggplot(lc_prng_df, aes(x = y)) +
  geom_histogram(fill = "#ffffe8", color="black", size=0.25,
                 boundary = 0, bins=16) +
  scale_x_continuous(lim = c(0, 2^16 - 1)) +
  ggtheme_tufte()
lc_prng_plot
```

That looks good, but as we suggested when we first introduced
pseudorandom number generators, we'll be able to do better than a
simple $\chi$-by-eye test.

## Converting to a $\mbox{uniform}(0, 1)$ sampler

We can now take our sampler for an integer range and divide by the
range to produce a continuous sampler.  To generate a continuous
sample from $\mathrm{uniform}(0, 1)$, we just generate a discrete
uniform draw from $0$ to $M$ and divide by $M$.^[Our pseudocode is
assuming we can store the seed value internally to the function and
that `M`, `alpha`, and `beta` are all fixed from the outside.]

```
uniform01_prng()
  seed = next_prng(seed, M, alpha, beta)
  return seed / M
```

Sometimes, the boundary values of zero and one are avoided by
returning `(seed + 1) / (M + 1)`.

## Mean and variance tests

One simple test for a pseudorandom number generator is whether it
produces the right means and variances.  We know that for a uniform
distribution between 0 and 1 that the mean is $\frac{1}{2}$ and its
variance is $\frac{1}{12}$.^[The mean follows by symmetry and the
variance is derived by solving $$\displaystyle \int_0^1 \left( y - \frac{1}{2} \right)^2 \,
\mathrm{d}y = \frac{1}{12}.$$]

Lets see what our mean and variances are from $10\,000$ draws from
our new function `uniform01_prng()`.

```{r}
y_unif <- y / M
printf("    sample mean = %3.3f", mean(y_unif))
printf("sample variance = %3.3f", var(y_unif))
```

Because $\frac{1}{12} \approx 0.083$, it looks like we're in the right
ballpark.  How do we turn these into proper tests, though?  We can't
just eybeall the results every time.

The central limit theorem gives us the ability to characterize the
behavior of sample averages, which is what both mean and variance
estimates are.^[Depending on the normality of the variables being
averaged, the approximation is often reasonable starting from as few
as ten draws and is usually very good with one hundred draws.]  For
example, in the simple case of the mean, we know that the variance of
a single draw is $\frac{1}{12}$, so the variance of the average of $N$
draws will be $\frac{1}{N \times 12}$, and the standard deviation is
$(N \times 12)^{-\frac{1}{2}} = \sqrt{\frac{1}{N \times 12}}$.

We can use the variance of the mean estimate to formulate a
probabilistic test that the mean is truly distributed as
$\mbox{normal}\left( 0.5, (N \times 12)^{-\frac{1}{2}} \right)$.  Thus we would
expect 99.9% of the draws to be within three standad deviations, or
$\pm 3 \times (N \times 12)^{-\frac{1}{2}}$.  With the $10\,000$ draws
we took, this is approximately 0.0029$, so it looks like the first run
of the mean test passed to within a single standard deviation.

## Probabilistic tests and false positives

We are unfortunately left with a test where we expect a false positive
failure one in a thousand runs.  That may be fine if we're only
running the test once or twice.  But if we're distributing this
software to thousands of people or regularly running our tests during
development, this is going to cause headaches.  The simple recourse we
have is to sharpen the test.  We expect the 99.997% of the draws to
fall within four standard deviations of the mean.  We might be able to
live with the resulting one in thirty thousand failure rate.  But we
may need to sharpen the test even further given our test conditions.
The problem with sharpening the test too far is that we can miss bugs
in less significant digits.

Another thing we can do for more power is run the test 100 times and
look at the distribution of means.  That distribution should itself be
normal and can be tested.  This is possible, and can lead to better
tests, but starts to get computationally prohibitive for large number
of draws per test.

The problem of requiring probabilistic tests for pobabilistic programs
pervades all of statistical software engineering.  There's no easy
solution.

## Binomial interval tests

If we are simulating a random variable $Y$, we know the probability that
it lies between $a$ and $b$ is given by^[Recall that we don't need to
worry about the points on either end, so can use less-than and
less-than-or-equal interchangeably in these formulas.]

$$
\begin{array}{rcl}
\mbox{Pr}[a < Y < b]
& = &
\displaystyle \int_a^b p_Y(y) \, \mathrm{d}y
\\[6pt]
& = &
F_Y(b) - F_Y(a).
\end{array}
$$

We can use this as the basis for a statistical test because it lets us
know the proportion of draws that are expected to fall in any given
interval.

Suppose we take $M$ draws $y^{(1)}, \ldots, y^{(M)}$ from our
pseudorandom number generator.  Our test statistic is the number of
draws $Z_{(a, b)}$ that fall in the interval $(a, b)$, i.e.,

$$
z_{(a, b)} = \sum_{m = 1}^M \mbox{I}[a < y^{(m)} < b].
$$

If the pseudorandom number generator is producing draws with the
proper distribution, then the test statistic is distributed as

$$
z_{(a, b)} \sim \mbox{binomial}(M, \mbox{Pr}[a < Y < b]).
$$

We'll perform a central test, rejecting the null hypothesis with
$p$-value $\alpha$ if the test statistic falls outside the central $1
- \alpha$ interval, i.e., if either

$$
F_{Z_{(a,b)}}(z_{(a,b)}) < \frac{\alpha}{2}
$$

or^[The second condition is rendered
parallel to the first with the complementary cumulative distribution
function, as $$F^{\complement}_{Z_{(a,b)}}(z_{(a,b)}) <
\frac{\alpha}{2}.$$]

$$
F_{Z_{(a,b)}}(z_{(a,b)}) > 1 - \frac{\alpha}{2}.
$$

To illustrate how this works, let's take our uniform pseudorandom
number generator and test it on the interval $(0, 0.5)$.^[50%
intervals are convenient for testing---more extreme probabilities lead
to smaller expected counts either inside or outside the interval.]

Suppose we are given the putatively uniformly distributed values
$y^{(1)}, \ldots, y^{(M)}$ for testing.  We can write a simple program
to generate a p-value for the central hypothesis test as follows.^[In
general, the ternary *conditional operator* is defined so that `cond ?
a : b` evaluates to `a` if `cond` evaluates to true, and to `b`
otherwise.]

```
m = sum(y < 0.5)
a = binomial_cdf(m | M, 0.5)
print 'm = ' m ' out of M = ' M
print 'reject uniformity with p-value '
      (a < 0.5) ? a : (1 - a)
```

The first line counts the number of draws $y^{(m)}$ for which $y^{(m)}
< 0.5$.  The second line computes the cumulative distribution function
value.  The third line manages the tail issues---if the value is less
than one half it's in the lower tail and we return it as is; otherwise
it is in the upper tail and we return the distance from one.

Let's see what happens when we run it.

```{r}
M <- 10000
u <- y / 2^16
m <- sum(u < 0.5)
a <- pbinom(m, M, 0.5)
printf('m = %d out of M = %d', m, M)
printf('reject unformity with p-value %3.2f',
       ifelse(a < 0.5, a, 1 - a))
```

So our smapler passes this simple test.  Ideally we'd want to test it
with a range of seeds to make sure passing isn't seed dependent as it
may be in these simple pseudorandom number generators.


## Chi-squared interval tests

A more fine-grained interval test that can handle multiple intervals
simultaneously relies on a normal approximation to the binomial.  We
know that as $N$ grows, $\mbox{binomial}(N, \theta)$ approaches a
normal distribution.^[The normal distribution it approaches has the
same mean and standard deviation, i.e., $\mbox{binomial}(N, \theta)$
is well approximated by $\mbox{normal}(N \times \theta, \sqrt{N \times
\theta \times (1 - \theta)})$ as $N$ grows.]

The chi-squared test is going to require the domain of the variable
being tested to be divided into $K$ exclusive and exhausive regions.
We'll assume that's done by dividing the domain into intervals with
dividing points points $a_1, \ldots, a_{K-1}$, producing the
folllowing intervals.^[As before, we needn't worry about points on the
boundaries as they have probability zero.]

$$
\begin{array}{rcc}
A_1 & = & (-\infty, a_1)
\\
A_2 & = & (a_1, a_2)
\\
\vdots & \vdots & \vdots
\\
A_K & = & (a_{K-1}, \infty)
\end{array}
$$

Now we let $z_k$ be the count of the number of draws falling in the
interval $A_k$.  The expected number of draws falling into interval
$A_k$ will be defined as

$$
E_k = M \times \mbox{Pr}[Y \in A_k].
$$

We are going to base the test on the following test statistic.^[This
was first introduced by Karl Pearson in Pearson, K., 1900. X. On the
criterion that a given system of deviations from the probable in the
case of a correlated system of variables is such that it can be
reasonably supposed to have arisen from random sampling. *The London,
Edinburgh, and Dublin Philosophical Magazine and Journal of Science*,
50(302):157--175.]

$$
X^2
\ = \
\sum_{k=1}^K
\frac{\left( z_k - E_k \right)^2}
     {E_k}.
$$

Because the $E_k$ is a constant and $z_k$ is approximately normal,
$z_k - E_k$ will also be approximately normal, and thus $\left( z_k -
E_k \right)^2$ will be roughly a squared normal, so that the sum of
$K$ such terms will be roughy a chi-squared distribution with $K - 1$
degrees of freedom.^[This follows from the definition of the
chi-squared distribution. Specifically, if $$Y_1, \ldots, Y_N \sim
\mbox{normal}(0, 1),$$ then $$(Y_1^2 + \cdots + Y_K^2) \sim
\mbox{chi\_squared}(K - 1)$$ has a *chi-squared distribution* with $K
- 1$ degrees of freedom.  The subtraction of $E_k$ and division by
$\sqrt{E_k}$ normalizes the distribution so that approximately
$$\frac{z_k - E_k}{\sqrt{E_k}} \sim \mbox{normal}(0, 1),$$ with the
approximation being optimal when the probabilty of success on which
the draws are based is 0.5.]  That means that if the null hypothesis
of the distribution of pseudorandom numbers being uniform is correct,
then we can set up the same kind of central hypothesis test as for the
ibnomial test.  Specifically, we'll reject the null hypothesis that
the pseudorandom number generator is behaving properly at significance
level $\alpha$ if

$$
\mbox{chi\_squared\_cdf}(X_2, K - 1) < \frac{\alpha}{2}
$$

or if

$$
\mbox{chi\_squared\_cdf}(X_2, K - 1) > 1 - \frac{\alpha}{2}.
$$

That is, we reject the null hypothesis at signifiance level $\alpha$
if the test statistic $X^2$ is outside of the central $1 - \alpha$
interval of the chi-squared distribution.

There are lots of choices when using the chi-squared test, like how
many bins to use and how to space them.  Generally, we want the
expected number of elements in each bin to be good enough that the
normal approximation is approximation, the traditionally suggested
minimum for which is five or so.  If we're testing a pseudorandom
number generator, the only bound is compute time, so we'll usually
have many more expected elements per bin than five.  It's common to
see equal probability bins used, generating them with an inverse
cumulative distribution function where available.
